---
title: "`r params$title`"
subtitle: "Partial Least Squares Analysis"
author: "`r params$author`"
date: "`r Sys.Date()`"
format:
  html:
    theme:
      light: [flatly, custom.scss]
      dark: [darkly, custom.scss]
    toc: true
    toc-location: left
    toc-depth: 3
    toc-title: "Contents"
    code-fold: true
    code-tools: true
    self-contained: true
    fig-width: 10
    fig-height: 6
    fig-dpi: 150
    smooth-scroll: true
    anchor-sections: true
    fontsize: 1rem
    linestretch: 1.5
    mainfont: "Inter, system-ui, sans-serif"
    max-width: 860px
include-in-header:
  text: |
    <style>
    .callout { margin: 24px 0; }
    .quarto-title { margin-bottom: 8px; }
    .quarto-title .subtitle { font-weight: 400; color: var(--bs-secondary-color); }
    h1, h2, h3 { font-weight: 600; letter-spacing: -0.02em; }
    .panel-tabset .nav-link { font-size: 14px; }
    table { font-size: 14px; }
    .stat-card { background: var(--bs-tertiary-bg); border-radius: 8px; padding: 24px 16px; text-align: center; border: 0.5px solid var(--bs-border-color); }
    .stat-value { font-size: 1.75rem; font-weight: 700; line-height: 1.2; margin-bottom: 4px; font-variant-numeric: tabular-nums; }
    .stat-label { font-size: 13px; color: var(--bs-secondary-color); font-weight: 500; }
    .status-dot { display: inline-block; width: 6px; height: 6px; border-radius: 50%; margin-right: 6px; }
    .status-dot.success { background: #22c55e; }
    .status-dot.neutral { background: var(--bs-secondary-color); }
    </style>
params:
  result_file: ""
  context_file: ""
  title: "PLS Analysis Report"
  author: ""
  include_brain: true
  bsr_threshold: 3
  p_threshold: 0.05
  method: 1
  n_lv: 5
  has_perm: true
  has_boot: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(plsrri)
library(ggplot2)

# Load result
result <- readRDS(params$result_file)
context <- if (nzchar(params$context_file) && file.exists(params$context_file)) {
  readRDS(params$context_file)
} else {
  list()
}

or_else <- function(x, y) {
  if (is.null(x) || (is.character(x) && length(x) == 1L && !nzchar(x))) y else x
}

group_subject_counts <- function(x) {
  if (is.null(x) || !length(x)) return(integer(0))
  vapply(x, function(group_counts) {
    if (length(group_counts) == 1L && !is.list(group_counts)) {
      return(as.integer(group_counts))
    }
    vals <- as.integer(unlist(group_counts, recursive = TRUE, use.names = FALSE))
    if (!length(vals)) return(0L)
    max(vals, na.rm = TRUE)
  }, integer(1))
}

subject_counts <- group_subject_counts(result$num_subj_lst)
subject_total <- sum(subject_counts)
group_count_text <- if (length(subject_counts)) {
  paste(subject_counts, collapse = " + ")
} else {
  "—"
}
feature_count <- nrow(result$u)

emit_plot <- function(expr, note) {
  plot_obj <- tryCatch(
    eval.parent(substitute(expr)),
    error = function(e) e
  )

  if (inherits(plot_obj, "error")) {
    cat("::: {.callout-note appearance='simple'}\n")
    cat(note, "\n\n")
    cat("Reason: `", conditionMessage(plot_obj), "`\n", sep = "")
    cat(":::\n\n")
    return(invisible(NULL))
  }

  rendered <- tryCatch(
    {
      print(plot_obj)
      NULL
    },
    error = function(e) e
  )

  if (inherits(rendered, "error")) {
    cat("::: {.callout-note appearance='simple'}\n")
    cat(note, "\n\n")
    cat("Reason: `", conditionMessage(rendered), "`\n", sep = "")
    cat(":::\n\n")
    return(invisible(NULL))
  }

  cat("\n\n")
  invisible(plot_obj)
}

# Helper for method names
method_name <- plsrri:::PLS_METHOD_NAMES[result$method]

# Count significant LVs
if (params$has_perm) {
  n_sig <- sum(result$perm_result$sprob < params$p_threshold)
  sig_lvs <- which(result$perm_result$sprob < params$p_threshold)
} else {
  n_sig <- NA
  sig_lvs <- seq_len(min(3, length(result$s)))
}
```

## Overview {.unnumbered}

::: {.grid}

::: {.g-col-12 .g-col-md-3}
::: {.stat-card}
::: {.stat-value}
`r method_name`
:::
::: {.stat-label}
Method
:::
:::
:::

::: {.g-col-6 .g-col-md-3}
::: {.stat-card}
::: {.stat-value}
`r length(result$s)`
:::
::: {.stat-label}
Latent variables
:::
:::
:::

::: {.g-col-6 .g-col-md-3}
::: {.stat-card}
::: {.stat-value}
`r if(!is.na(n_sig)) n_sig else "—"`
:::
::: {.stat-label}
Significant (p < `r params$p_threshold`)
:::
:::
:::

::: {.g-col-6 .g-col-md-3}
::: {.stat-card}
::: {.stat-value}
`r subject_total`
:::
::: {.stat-label}
Subjects
:::
:::
:::

:::

---

## Parameters

| Parameter | Value |
|:----------|:------|
| Groups | `r length(subject_counts)` (`r group_count_text` subjects) |
| Conditions | `r result$num_cond` |
| Features | `r format(feature_count, big.mark = ",")` |
| Permutations | `r if(params$has_perm) format(result$perm_result$num_perm, big.mark = ",") else "—"` |
| Bootstrap samples | `r if(params$has_boot) format(result$boot_result$num_boot, big.mark = ",") else "—"` |

## Provenance

```{r}
#| label: provenance
#| output: asis

meta <- or_else(context$metadata, list())
pipeline_summary <- context$pipeline_summary

if (!length(meta) && is.null(pipeline_summary)) {
  cat("::: {.callout-note appearance='simple'}\n")
  cat("This report was rendered from an in-memory or standalone result object. Pipeline provenance was not available.\n")
  cat(":::\n")
} else {
  artifact_root <- or_else(meta$artifact_root, "—")
  spec_path <- or_else(meta$spec_path, "—")
  generated_at <- or_else(meta$generated_at, "—")
  input_kind <- or_else(meta$input_kind, "—")
  package_version <- or_else(meta$package_version, as.character(packageVersion("plsrri")))

  cat("| Field | Value |\n")
  cat("|:------|:------|\n")
  cat("| Input type | ", input_kind, " |\n", sep = "")
  cat("| Artifact root | `", artifact_root, "` |\n", sep = "")
  cat("| Pipeline spec | `", spec_path, "` |\n", sep = "")
  cat("| Generated at | ", generated_at, " |\n", sep = "")
  cat("| plsrri version | ", package_version, " |\n", sep = "")
}
```

```{r}
#| label: pipeline-summary

if (!is.null(context$pipeline_summary)) {
  print(knitr::kable(context$pipeline_summary, align = rep("l", ncol(context$pipeline_summary))))
}
```

## Input Inventory

```{r}
#| label: input-inventory
#| output: asis

discovery <- context$discovery
pls_manifest <- context$pls_manifest

if (is.null(discovery) && is.null(pls_manifest)) {
  cat("::: {.callout-note appearance='simple'}\n")
  cat("No discovery or planned-input inventory was saved with this result.\n")
  cat(":::\n")
} else {
  if (!is.null(discovery)) {
    cat("### Discovery Manifest\n\n")
    preview <- utils::head(discovery, 10)
    print(knitr::kable(preview, align = rep("l", ncol(preview))))
    cat("\n\n")
  }

  if (!is.null(pls_manifest)) {
    cat("### PLS Manifest\n\n")
    preview <- utils::head(pls_manifest, 12)
    print(knitr::kable(preview, align = rep("l", ncol(preview))))
    cat("\n\n")
  }
}
```

## Singular Values

Singular values quantify the strength of each latent variable. Larger values indicate patterns that explain more cross-block covariance.

```{r}
#| label: sv-plot
#| fig-height: 4.5

emit_plot(
  plot_singular_values(result, alpha = params$p_threshold) + labs(title = NULL),
  "Singular value visualization was not available for this saved result."
)
```

```{r}
#| label: sv-table

df <- data.frame(
  LV = seq_along(result$s),
  `Singular Value` = round(result$s, 4),
  `Variance (%)` = round((result$s^2) / sum(result$s^2) * 100, 1),
  `Cumulative (%)` = round(cumsum((result$s^2) / sum(result$s^2) * 100), 1),
  check.names = FALSE
)

if (params$has_perm) {
  df$`P-value` <- format(round(result$perm_result$sprob, 4), nsmall = 4)
  # Status dot pattern: green dot for significant
  df$Status <- ifelse(
    result$perm_result$sprob < params$p_threshold,
    '<span class="status-dot success"></span>Significant',
    ""
  )
}

knitr::kable(df, align = c("c", "r", "r", "r", "r", "l"), escape = FALSE)
```

## Latent Variables

```{r}
#| label: sig-status
#| output: asis

if (params$has_perm) {
  if (length(sig_lvs) == 0) {
    cat("::: {.callout-warning}\n")
    cat("No latent variables reached significance at p <", params$p_threshold, ".\n")
    cat("Showing top 3 LVs for exploratory purposes.\n")
    cat(":::\n\n")
    sig_lvs <- seq_len(min(3, length(result$s)))
  } else {
    cat("::: {.callout-tip appearance='simple'}\n")
    cat('<span class="status-dot success"></span>')
    cat(length(sig_lvs), "latent variable(s) reached significance: LV",
        paste(sig_lvs, collapse = ", LV"), "\n")
    cat(":::\n\n")
  }
}
```

::: {.panel-tabset}

```{r}
#| label: lv-tabs
#| output: asis
#| fig-height: 5

for (lv in sig_lvs[seq_len(min(params$n_lv, length(sig_lvs)))]) {

  # Tab header - dot indicates significance
  var_exp <- round((result$s[lv]^2) / sum(result$s^2) * 100, 1)
  is_sig <- params$has_perm && result$perm_result$sprob[lv] < params$p_threshold

  cat("\n\n### LV", lv, "\n\n")

  # Stats summary - flat grid, no decoration
  cat("::: {.grid}\n\n")

  cat("::: {.g-col-4}\n")
  cat("**Variance**\n\n")
  cat(var_exp, "%\n")
  cat(":::\n\n")

  if (params$has_perm) {
    p_val <- result$perm_result$sprob[lv]
    cat("::: {.g-col-4}\n")
    cat("**P-value**\n\n")
    if (is_sig) {
      cat('<span class="status-dot success"></span>')
    }
    cat(format(round(p_val, 4), nsmall = 4), "\n")
    cat(":::\n\n")
  }

  if (params$has_boot) {
    bsr_vals <- bsr(result, lv = lv)
    n_reliable <- sum(abs(bsr_vals) > params$bsr_threshold)
    cat("::: {.g-col-4}\n")
    cat("**Reliable voxels**\n\n")
    cat(format(n_reliable, big.mark = ","), " (|BSR| > ", params$bsr_threshold, ")\n", sep = "")
    cat(":::\n\n")
  }

  cat(":::\n\n")

  # Design scores
  cat("#### Design Scores\n\n")
  cat("Design scores show how strongly each condition-group expresses this brain pattern.\n\n")
  emit_plot(
    plot_scores(result, lv = lv, type = "design", plot_type = "bar") +
      theme(plot.title = element_blank()),
    "Design score plotting was not available for this latent variable."
  )

  # Loadings
  cat("#### Loadings\n\n")
  emit_plot(
    plot_loadings(result, lv = lv, type = "design") +
      theme(plot.title = element_blank()),
    "Loading visualization was not available for this latent variable."
  )

  # Brain scores
  cat("#### Brain Scores\n\n")
  cat("Individual brain scores. Higher values indicate stronger expression of the pattern.\n\n")
  emit_plot(
    plot_scores(result, lv = lv, type = "brain", plot_type = "violin") +
      theme(plot.title = element_blank()),
    "Brain score plotting was not available for this latent variable."
  )
}
```

:::

## Bootstrap Reliability

```{r}
#| label: bootstrap-section
#| output: asis

if (params$has_boot) {
  cat("Bootstrap resampling (n = ", format(result$boot_result$num_boot, big.mark = ","),
      ") assesses the reliability of voxel contributions.\n\n", sep = "")
  cat("**Bootstrap Ratio (BSR)** = salience / standard error\n\n")
  cat("Voxels with |BSR| > ", params$bsr_threshold,
      " are considered reliably contributing to the pattern.\n\n", sep = "")
} else {
  cat("::: {.callout-note appearance='simple'}\n")
  cat("Bootstrap analysis was not performed. Run with `nboot > 0` to assess voxel reliability.\n")
  cat(":::\n")
}
```

```{r}
#| label: bootstrap-summary
#| eval: !expr params$has_boot

boot_summary <- data.frame(
  LV = sig_lvs[seq_len(min(5, length(sig_lvs)))],
  stringsAsFactors = FALSE
)

boot_summary$`Reliable Voxels` <- sapply(boot_summary$LV, function(lv) {
  bsr_vals <- bsr(result, lv = lv)
  sum(abs(bsr_vals) > params$bsr_threshold)
})

boot_summary$`% of Total` <- round(boot_summary$`Reliable Voxels` / feature_count * 100, 2)

boot_summary$`Max +BSR` <- sapply(boot_summary$LV, function(lv) {
  round(max(bsr(result, lv = lv)), 2)
})

boot_summary$`Max -BSR` <- sapply(boot_summary$LV, function(lv) {
  round(min(bsr(result, lv = lv)), 2)
})

knitr::kable(boot_summary, align = c("c", "r", "r", "r", "r"))
```

## Artifact Notes

```{r}
#| label: artifact-notes
#| output: asis

validation <- context$validation
pls_summary <- context$pls_summary

if (!is.null(validation) && nrow(validation) > 0) {
  cat("### Validation\n\n")
  print(knitr::kable(validation, align = rep("l", ncol(validation))))
  cat("\n\n")
}

if (!is.null(pls_summary) && nrow(pls_summary) > 0) {
  cat("### Saved Summary\n\n")
  print(knitr::kable(pls_summary, align = rep("l", ncol(pls_summary))))
  cat("\n\n")
}
```

## Methods

::: {.callout-note appearance="minimal" collapse="true"}
## Methodological details

### Analysis Type

**`r method_name`**

```{r}
#| output: asis

method_desc <- switch(as.character(result$method),
  "1" = "Mean-centering Task PLS identifies distributed brain patterns that optimally differentiate experimental conditions. Data are mean-centered within groups, and singular value decomposition extracts orthogonal latent variables that maximize the covariance between conditions and brain activity.",
  "2" = "Non-rotated Task PLS tests specific a priori hypotheses about condition differences using predefined design contrasts. Unlike rotated PLS, the solution is constrained to the contrast space.",
  "3" = "Behavior PLS identifies brain patterns that maximally correlate with behavioral measures. The analysis extracts latent variables that capture the strongest brain-behavior relationships across conditions.",
  "4" = "Multiblock PLS combines task effects and behavior-brain correlations in a unified analysis, partitioning variance into task-related and behavior-related components.",
  "5" = "Non-rotated Behavior PLS tests specific hypotheses about brain-behavior relationships using predefined contrasts on the behavioral variables.",
  "6" = "Non-rotated Multiblock PLS tests combined hypotheses about both task effects and behavior-brain relationships using design contrasts."
)

cat(method_desc)
```

### Statistical Inference

```{r}
#| output: asis

if (params$has_perm) {
  cat("**Permutation Testing:** ", format(result$perm_result$num_perm, big.mark = ","),
      " permutations were used to generate a null distribution of singular values. ",
      "P-values represent the proportion of permuted singular values exceeding the observed value.\n\n", sep = "")
}

if (params$has_boot) {
  cat("**Bootstrap Resampling:** ", format(result$boot_result$num_boot, big.mark = ","),
      " bootstrap samples were drawn (with replacement, stratified by group) to estimate ",
      "standard errors of voxel saliences. Bootstrap ratios (BSR = salience/SE) approximate ",
      "z-scores under the assumption of normally distributed saliences.\n\n", sep = "")
}
```

### References

- McIntosh, A. R., & Lobaugh, N. J. (2004). Partial least squares analysis of neuroimaging data: Applications and advances. *NeuroImage*, 23, S250-S263.
- Krishnan, A., Williams, L. J., McIntosh, A. R., & Hervé, A. (2011). Partial Least Squares (PLS) methods for neuroimaging. *NeuroImage*, 56(2), 455-475.

:::

## Session

::: {.callout-note appearance="minimal" collapse="true"}
## Technical details

```{r}
#| code-fold: false

sessionInfo()
```

:::

---

::: {style="text-align: center; font-size: 13px; color: var(--bs-secondary-color); margin-top: 32px;"}
Generated with [plsrri](https://github.com/bbuchsbaum/plsrri) v`r packageVersion("plsrri")` on `r format(Sys.time(), "%B %d, %Y")`
:::
