---
title: "Formula-Aware ASCA-PLS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Formula-Aware ASCA-PLS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

ASCA-PLS adds a formula-aware layer to Task PLS. The goal is to make the
design side behave like an ANOVA-style decomposition without pretending that
residualized test components are ordinary factorial effects.

The rule is:

> Inference is partial. Interpretation is factorial. The bridge is explicit.

Each formula term carries three linked objects:

- an **estimand**, the named factorial effect such as `group:task:level`;
- a **test**, the lower-order-residualized partial space used for evidence;
- a **display view**, usually `tested_effect`, which lifts the tested partial
  direction back into named factorial cell coordinates.

## Basic Workflow

```{r example, eval = FALSE}
fit <- asca_pls(
  spec,
  decompose = ~ group * task * level,
  condition_key = condition_key,
  id = "subject",
  within = c("task", "level"),
  between = "group",
  test = partial_test(
    method = "freedman_lane",
    statistic = "trace",
    nperm = 1000,
    correction = "maxT"
  ),
  selection = hierarchical_selection(
    method = "backward",
    heredity = "strong",
    alpha = 0.05
  )
)
```

`anova(fit)` returns the partial evidence table. `plot(fit, term = ...)`
defaults to the tested-effect display: the tested partial component chooses the
brain direction, while the design profile is shown in the original factorial
cell coordinates.

```{r accessors, eval = FALSE}
anova(fit)
selected_formula(fit)
asca_term_table(fit)

asca_components(fit, term = "group:task:level")
brain_saliences(fit, term = "group:task:level")
design_saliences(fit, term = "group:task:level")

plot_asca_effect_profile(
  fit,
  term = "group:task:level",
  x_axis = "level",
  line = "task",
  facet = "group"
)
plot_asca_alignment(fit)
plot_asca_selection(fit)
```

## Views

`view = "partial"` is the raw lower-order-residualized evidence space. It is
the right space for nested tests, but it is not the easiest scientific object
to interpret.

`view = "effect"` is the named factorial effect-space SVD. It is readable, but
its component orientation is descriptive rather than necessarily the component
used by the partial test.

`view = "tested_effect"` is the default compromise. It uses the partial
component for orientation and displays the corresponding profile after lifting
it into the original term coordinates.

## Current Permutation Scope

Reduced-model permutation inference requires the original `pls_spec`, even when
you also pass a precomputed `fit`. The implementation uses raw-observation
Freedman-Lane residual permutations and then rebuilds the Task PLS cross-block
for each permutation. For repeated-measures terms, residuals are permuted within
derived subject blocks. For pure between-subject terms, residuals are permuted
as whole subject blocks while preserving the condition-major row order expected
by Task PLS.

This distinction matters. A purely cell-level residual permutation can be
anti-conservative when the data contain only main effects plus subject-level
repeated-measures structure. The synthetic calibration tests therefore exercise
main-effects-only multivariate data and check that higher-order interaction and
between-subject null terms do not exceed binomial false-positive-rate bounds.

For neuroimaging interpretation, use permutation p-values for term/component
evidence and bootstrap-style stability summaries for feature saliences.
