---
title: "Behavior PLS Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Behavior PLS Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)
```

```{r load-package}
library(plsrri)
```

Behavior PLS finds latent variables that maximally capture the covariance
between brain features and continuous behavioral measures — reaction time,
accuracy, symptom scores, or any per-subject-condition variable. Where Task
PLS asks "which brain patterns separate conditions?", Behavior PLS asks
"which brain patterns track individual differences in behavior?"

## What data shape does Behavior PLS expect?

You need two matrices with the same stacked row structure (subjects within
conditions, then groups):

- **Brain matrix** — rows = observations, columns = voxels/features
- **Behavior matrix** — rows = observations, columns = behavioral measures

```{r simulate-behavior-data, include = FALSE}
set.seed(123)

n_subj <- 18
n_cond <- 2
n_feat <- 120
total_rows <- n_subj * n_cond

latent <- as.numeric(scale(rnorm(total_rows)))
condition_shift <- rep(c(-1, 1), each = n_subj)

behav_data <- cbind(
  rt = as.numeric(scale(1.4 * latent + 0.5 * condition_shift + rnorm(total_rows, sd = 0.35))),
  acc = as.numeric(scale(-1.1 * latent + 0.3 * condition_shift + rnorm(total_rows, sd = 0.35)))
)

brain_data <- matrix(rnorm(total_rows * n_feat, sd = 0.7), nrow = total_rows, ncol = n_feat)
brain_data[, 1:35] <- brain_data[, 1:35] + latent %o% rep(1.2, 35)
brain_data[, 36:70] <- brain_data[, 36:70] + latent %o% rep(-0.9, 35)
brain_data[, 71:90] <- brain_data[, 71:90] + condition_shift %o% rep(0.6, 20)
colnames(behav_data) <- c("rt", "acc")
```

This example has `r n_subj` subjects, `r n_cond` conditions, `r n_feat`
voxels, and two behavioral measures. The synthetic data has a single latent
factor driving both RT and accuracy in opposite directions, plus a
condition-specific voxel block:

```{r inspect-behavior-data}
head(round(behav_data, 2))
```

## How do you run a behavior analysis?

`behav_pls()` is the convenience wrapper. It runs Behavior PLS (method 3)
with the behavior matrix as the second block:

```{r run-behavior-pls}
behav_result <- behav_pls(
  datamat_lst = list(brain_data),
  behav_data = behav_data,
  num_subj_lst = n_subj,
  num_cond = n_cond,
  nperm = 200,
  nboot = 200,
  progress = FALSE
)
```

Check which latent variables are significant and how much variance each
explains:

```{r behavior-significance}
cbind(
  pvalue = round(significance(behav_result), 3),
  variance = round(singular_values(behav_result, normalize = TRUE), 1)
)
```

LV1 captures the dominant brain-behavior relationship. The scree plot
makes the structure clear:

```{r plot-behavior-scree, fig.cap = "Singular value scree for the behavior PLS example. LV1 dominates."}
plot_singular_values(behav_result)
```

## Which behaviors define the first latent variable?

In Behavior PLS, `loadings(..., type = "behavior")` returns the
correlation-style weights linking each behavior-by-condition cell to the
latent variable. Each row is one condition-measure combination.

```{r behavior-loadings}
round(loadings(behav_result, type = "behavior", lv = 1), 2)
```

RT and accuracy load with opposite signs, consistent with the planted
relationship — subjects who are faster also tend to be more accurate on the
latent dimension captured by LV1.

```{r plot-behavior-loadings, fig.cap = "Behavior loadings for LV1. RT and accuracy load in opposite directions, reflecting the shared latent factor."}
plot_loadings(behav_result, lv = 1, type = "behavior", plot_type = "dot")
```

## How strongly do subjects express that pattern?

Brain scores show how each observation projects onto the latent variable.
In a real analysis, you would check whether the pattern varies across
conditions or groups:

```{r plot-behavior-brain-scores, fig.cap = "Brain scores for LV1. The spread reflects individual differences in the brain-behavior relationship."}
plot_scores(behav_result, lv = 1, type = "brain", plot_type = "violin")
```

## Are the behavior correlations stable?

For behavior methods, `confidence(..., what = "correlation")` returns
bootstrap confidence intervals on the brain-behavior correlations. Intervals
that exclude zero indicate a reliable relationship:

```{r behavior-correlation-ci}
corr_ci <- confidence(behav_result, what = "correlation", lv = 1)
round(cbind(lower = corr_ci$lower[, 1], upper = corr_ci$upper[, 1]), 2)
```

Both measures show intervals that are clearly away from zero, confirming
the stability of the brain-behavior association under resampling.

## Which voxels contribute reliably?

Bootstrap ratios work the same way as in Task PLS — they indicate which
voxels participate reliably in the latent pattern:

```{r behavior-bsr}
behav_bsr <- bsr(behav_result, lv = 1)
reliable <- sum(abs(behav_bsr) > 2)
```

`r reliable` of `r n_features(behav_result)` voxels have $|BSR| > 2$.

```{r plot-behavior-bsr, echo = FALSE, fig.cap = "Bootstrap ratio profile. Voxels 1-35 and 36-70 show the planted latent factor; voxels 71-90 show the condition-specific signal.", fig.height = 3}
plot(as.numeric(behav_bsr), type = "h", col = "steelblue",
     xlab = "Voxel", ylab = "Bootstrap Ratio (LV1)",
     main = NULL, las = 1)
abline(h = c(-2, 2), lty = 2, col = "grey50")
```

The two voxel blocks (1-35 positive, 36-70 negative) mirror the planted
latent structure, while the condition block (71-90) contributes more weakly.

## Where to go next

| Goal | Resource |
|:-----|:---------|
| Task PLS workflow | `vignette("plsrri")` |
| Multiblock and seed PLS | `vignette("multiblock-and-seed")` |
| Behavior wrapper | `?behav_pls` |
| Full engine with all options | `?pls_analysis` |
| Loadings interpretation | `?plot_loadings` |
| Bootstrap confidence intervals | `?confidence` |
