---
title: "Command Line Interface"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Command Line Interface}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)

ensure_local_plsrri <- function(required_exports) {
  pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."
  if (!file.exists(file.path(pkg_root, "DESCRIPTION")) ||
      !requireNamespace("devtools", quietly = TRUE)) {
    return(invisible(FALSE))
  }

  installed_exports <- if (requireNamespace("plsrri", quietly = TRUE)) {
    getNamespaceExports("plsrri")
  } else {
    character(0)
  }

  if (!all(required_exports %in% installed_exports)) {
    devtools::load_all(pkg_root, quiet = TRUE)
    return(invisible(TRUE))
  }

  invisible(FALSE)
}

ensure_local_plsrri(c("install_cli", "plscli_main"))
```

```{r load-package}
library(plsrri)
```

`plsrri` ships an installable `plscli` wrapper for staged non-GUI workflows.
The command is designed for the same pipeline contract that powers
`prepare_firstlevel()`, `prepare_pls()`, `run_pls()`, and `render_pls_report()`.

This article is about the shell interface itself: installation, help, command
shapes, machine-readable output, and exit behavior. For the YAML schema, see
`vignette("pipeline-yaml-spec")`. For a broader end-to-end analysis example,
see `vignette("scripted-workflows")`.

## How do you install the `plscli` wrapper?

Install the package, then copy the wrapper into a directory on `PATH`.

```r
plsrri::install_cli("~/.local/bin", overwrite = TRUE)
```

If needed, add that directory to your shell startup file:

```bash
export PATH="$HOME/.local/bin:$PATH"
```

Then check the command:

```bash
plscli --help
plscli help report
```

When working from a source checkout, the wrapper can also bootstrap the local
package via `pkgload` or `devtools` if `plsrri` is not yet installed.

## What commands does `plscli` expose?

The command surface is intentionally small and stage-oriented:

- `template`
- `validate`
- `discover`
- `firstlevel-plan`
- `firstlevel-run`
- `pls-plan`
- `pls-run`
- `report`
- `summarize`
- `run`

The usual starting point is a scaffolded YAML file:

```bash
plscli template --out study.yml
```

Then validate the specification before launching expensive work:

```bash
plscli validate --spec study.yml
```

## What does a typical staged workflow look like?

For a workstation or scheduler-driven run, the stage boundaries are explicit:

```bash
plscli discover --spec study.yml
plscli firstlevel-plan --spec study.yml
plscli firstlevel-run --spec study.yml --work-id w0007
plscli pls-plan --spec study.yml
plscli pls-run --spec study.yml
plscli report --input study.yml --format html
```

For array execution, the main difference is the first-level worker id:

```bash
plscli firstlevel-plan --spec study.yml
plscli firstlevel-run --spec study.yml --work-id "${SLURM_ARRAY_TASK_ID}"
plscli pls-plan --spec study.yml
plscli pls-run --spec study.yml
```

The `report` command accepts either the YAML spec, an artifact root, or a saved
`pls_result.rds` path:

```bash
plscli report --input /path/to/plscli-out --format pdf --output report.pdf
```

## How do options and help behave?

Commands support both `--key value` and `--key=value` forms:

```bash
plscli report --input=study.yml --format=html
```

Command-specific help is available in either style:

```bash
plscli help firstlevel-run
plscli firstlevel-run --help
```

Unknown options are treated as usage errors rather than being ignored. Boolean
flags use `--flag`, and `report` also accepts `--no-open` when you want to
disable browser launching explicitly.

## Can the CLI emit machine-readable output?

Yes. Add `--json` when you want a stable stdout payload for scripts or CI:

```bash
plscli template --out study.yml --json
plscli validate --spec study.yml --json
plscli report --input study.yml --format html --json
```

Human-facing diagnostics still go to stderr where practical.

## What exit codes should you expect?

The CLI returns:

- `0` for success
- `1` for domain failures such as an invalid pipeline spec
- `2` for usage or runtime errors

That makes it safe to call from shell scripts, schedulers, and CI wrappers:

```bash
plscli validate --spec study.yml --json
status=$?
if [ "$status" -ne 0 ]; then
  echo "validation failed with status $status" >&2
  exit "$status"
fi
```

## Where should you go next?

- `vignette("pipeline-yaml-spec")` for the YAML contract
- `vignette("scripted-workflows")` for the broader saved-artifact workflow
- `?install_cli`
- `?plscli_main`
