---
title: "Group by Covariate Designs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Group by Covariate Designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)

ensure_local_plsrri <- function(required_exports) {
  pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."
  if (!file.exists(file.path(pkg_root, "DESCRIPTION")) ||
      !requireNamespace("devtools", quietly = TRUE)) {
    return(invisible(FALSE))
  }

  installed_exports <- if (requireNamespace("plsrri", quietly = TRUE)) {
    getNamespaceExports("plsrri")
  } else {
    character(0)
  }

  if (!all(required_exports %in% installed_exports)) {
    devtools::load_all(pkg_root, quiet = TRUE)
    return(invisible(TRUE))
  }

  invisible(FALSE)
}

ensure_local_plsrri(c("pls_spec", "add_subjects", "add_conditions",
                      "add_behavior", "add_design", "configure", "run"))
```

```{r load-package}
library(plsrri)
```

You have one brain map per subject, two between-subject groups, and a
trait-level continuous covariate. In formula notation, the motivating
model is often written as:

```text
brain map ~ group * trait
```

In `plsrri`, this is not fit as a voxelwise formula model. The PLS version
asks for multivariate brain patterns associated with the trait, and for
whether that trait-brain pattern is shared by the groups or differs between
them.

## What is the PLS translation?

With one map per subject, use a single condition:

- the two groups are the two entries of `datamat_lst`
- `num_cond = 1`
- the trait is a one-column behavior matrix
- the behavior matrix rows are stacked in the same order as the brain data

Behavior PLS computes a trait-brain association separately within each
group. With two groups, one condition, and one trait, the behavior side has
two rows:

```text
group 1 : trait-brain association
group 2 : trait-brain association
```

Contrasts over these two rows let you ask for the common trait association
and the group difference in that association.

## What data shape do we need?

Here is a compact simulated example. `Y_g1` and `Y_g2` are subject-by-feature
matrices. In a real analysis, the columns would be voxels, parcels, surface
vertices, or another common feature space.

```{r simulate-data}
set.seed(42)

n1 <- 28
n2 <- 20
n_features <- 120

trait_g1 <- as.numeric(scale(rnorm(n1)))
trait_g2 <- as.numeric(scale(rnorm(n2)))

Y_g1 <- matrix(rnorm(n1 * n_features, sd = 0.7), nrow = n1)
Y_g2 <- matrix(rnorm(n2 * n_features, sd = 0.7), nrow = n2)
```

```{r add-signal, include = FALSE}
shared_pattern <- c(rep(1.0, 30), rep(-0.8, 30), rep(0, n_features - 60))
interaction_pattern <- c(rep(0, 60), rep(1.1, 25), rep(-0.9, 25),
                         rep(0, n_features - 110))
group_pattern <- c(rep(0, 90), rep(0.8, 30))

Y_g1 <- Y_g1 + trait_g1 %o% shared_pattern - 0.35 * trait_g1 %o% interaction_pattern
Y_g2 <- Y_g2 + trait_g2 %o% shared_pattern + 0.35 * trait_g2 %o% interaction_pattern
Y_g2 <- sweep(Y_g2, 2, group_pattern, "+")

trait <- matrix(c(trait_g1, trait_g2), ncol = 1)
colnames(trait) <- "trait"

stopifnot(
  nrow(Y_g1) == n1,
  nrow(Y_g2) == n2,
  nrow(trait) == n1 + n2,
  all(is.finite(Y_g1)),
  all(is.finite(Y_g2)),
  all(is.finite(trait))
)
```

The trait matrix is stacked as group 1 followed by group 2:

```{r inspect-trait}
head(trait, 4)
tail(trait, 4)
```

## How do we test the trait effect and the interaction?

Use non-rotated Behavior PLS when the hypothesis is explicit. The design
matrix has one row per group-specific trait association. The first column
asks for a shared trait-brain association; the second asks whether the
association differs by group.

```{r behavior-contrasts}
trait_design <- cbind(
  common_trait = c(1, 1),
  group_by_trait = c(-1, 1)
)

trait_design
```

```{r fit-behavior-nonrotated}
trait_result <- pls_spec() |>
  add_subjects(list(Y_g1, Y_g2), groups = c(n1, n2)) |>
  add_conditions(1) |>
  add_behavior(trait) |>
  add_design(trait_design) |>
  configure(
    method = "behavior_nonrotated",
    cormode = "pearson",
    nperm = 199,
    nboot = 199
  ) |>
  run(progress = FALSE)
```

```{r check-trait-result, include = FALSE}
trait_p <- significance(trait_result)
trait_var <- singular_values(trait_result, normalize = TRUE)
stopifnot(
  inherits(trait_result, "pls_result"),
  length(trait_p) == 2,
  all(is.finite(trait_p)),
  all(trait_p >= 0 & trait_p <= 1),
  all(is.finite(trait_var))
)
```

The two requested contrasts become two latent variables:

```{r trait-summary}
cbind(
  pvalue = round(significance(trait_result), 3),
  variance = round(singular_values(trait_result, normalize = TRUE), 1)
)
```

The design loadings show the hypothesis weights used for each latent
variable:

```{r trait-loadings}
loadings(trait_result, type = "design")
```

Interpret the rows as:

- `common_trait`: the trait-brain association pooled across groups
- `group_by_trait`: the difference in trait-brain association between groups

The sign of `group_by_trait` follows the contrast coding. With `c(-1, 1)`,
positive brain saliences are more strongly associated with the trait in
group 2 than in group 1.

## Where is the group main effect?

The trait model above targets the covariate part of `group * trait`. It
does not estimate the mean group difference. For the main group effect, use
non-rotated Task PLS with one condition and a group contrast:

```{r fit-group-main}
group_result <- pls_spec() |>
  add_subjects(list(Y_g1, Y_g2), groups = c(n1, n2)) |>
  add_conditions(1) |>
  add_design(cbind(group_2_minus_1 = c(-1, 1))) |>
  configure(
    method = "task_nonrotated",
    nperm = 199,
    nboot = 199
  ) |>
  run(progress = FALSE)
```

```{r check-group-result, include = FALSE}
group_p <- significance(group_result)
group_var <- singular_values(group_result, normalize = TRUE)
stopifnot(
  inherits(group_result, "pls_result"),
  length(group_p) == 1,
  all(is.finite(group_p)),
  all(group_p >= 0 & group_p <= 1),
  all(is.finite(group_var))
)
```

```{r group-summary}
cbind(
  pvalue = round(significance(group_result), 3),
  variance = round(singular_values(group_result, normalize = TRUE), 1)
)
```

This is usually the clearest workflow: fit the group mean effect and the
trait interaction question as separate, targeted analyses.

## Can this be one multiblock analysis?

Yes. Non-rotated Multiblock PLS combines the task block and behavior block.
For this design, the multiblock design rows are:

```text
group 1 mean
group 2 mean
group 1 trait association
group 2 trait association
```

That gives a single analysis with columns for the group main effect, common
trait effect, and group-by-trait effect.

```{r multiblock-design}
multiblock_design <- cbind(
  group_2_minus_1 = c(-1, 1, 0, 0),
  common_trait = c(0, 0, 1, 1),
  group_by_trait = c(0, 0, -1, 1)
)

multiblock_design
```

```{r fit-multiblock}
multiblock_result <- pls_spec() |>
  add_subjects(list(Y_g1, Y_g2), groups = c(n1, n2)) |>
  add_conditions(1) |>
  add_behavior(trait) |>
  add_design(multiblock_design) |>
  configure(
    method = "multiblock_nonrotated",
    cormode = "pearson",
    nperm = 199,
    nboot = 199
  ) |>
  run(progress = FALSE)
```

```{r check-multiblock-result, include = FALSE}
multiblock_p <- significance(multiblock_result)
multiblock_var <- singular_values(multiblock_result, normalize = TRUE)
stopifnot(
  inherits(multiblock_result, "pls_result"),
  length(multiblock_p) == 3,
  all(is.finite(multiblock_p)),
  all(multiblock_p >= 0 & multiblock_p <= 1),
  all(is.finite(multiblock_var))
)
```

```{r multiblock-summary}
cbind(
  pvalue = round(significance(multiblock_result), 3),
  variance = round(singular_values(multiblock_result, normalize = TRUE), 1)
)
```

Use the multiblock version when you want the mean group difference and the
trait associations represented in one decomposition. Use the separate
non-rotated analyses when you want the most direct test of each term.

## What should you report?

For the trait analysis, report:

- the row order and contrast coding
- whether the covariate was centered or scaled before analysis
- the permutation p-values for the requested LVs
- the bootstrap ratios or confidence intervals for stable brain features
- the sign convention for the interaction contrast

For unbalanced groups, pass the actual group sizes in `groups = c(n1, n2)`.
The bootstrap and permutation routines use those group sizes when resampling.

## Where to go next

| Goal | Resource |
|:-----|:---------|
| Behavior PLS with continuous measures | `vignette("behavior-pls")` |
| Group and condition effects | `vignette("plsrri")` |
| Multiblock PLS | `vignette("multiblock-and-seed")` |
| Predictive validation from PLS scores | `vignette("predictive-measures")` |
| Full engine options | `?pls_analysis` |
