---
title: "Pipeline YAML Specification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pipeline YAML Specification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)

needed <- c("yaml")
missing_pkgs <- needed[!vapply(needed, requireNamespace, quietly = TRUE, FUN.VALUE = logical(1))]
if (length(missing_pkgs) > 0) {
  knitr::knit_exit()
}

pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."
if (requireNamespace("plsrri", quietly = TRUE) &&
    !"write_pipeline_template" %in% getNamespaceExports("plsrri") &&
    requireNamespace("devtools", quietly = TRUE) &&
    file.exists(file.path(pkg_root, "DESCRIPTION"))) {
  devtools::load_all(pkg_root, quiet = TRUE)
}
```

```{r load-package}
library(plsrri)
library(yaml)
```

The YAML file is the durable contract for the non-GUI pipeline. It is what
`prepare_firstlevel()`, `prepare_pls()`, `run_pls()`, `plscli`, and the Shiny
import/export path all share.

This article is a reference for that contract. It focuses on structure,
required fields, and common variants. For the scripted workflow itself, see
`vignette("scripted-workflows")`.

## What sections does the spec contain?

The scaffold written by `write_pipeline_template()` gives the intended shape.

```{r template-scaffold, include = FALSE}
template_file <- tempfile(fileext = ".yml")
write_pipeline_template(template_file)
template_spec <- read_pipeline_spec(template_file)

required_sections <- c("dataset", "design", "first_level", "pls", "execution", "outputs")
stopifnot(all(required_sections %in% names(template_spec)))
```

```{r template-keys}
required_sections
```

Those sections have distinct roles:

- `dataset`: where the BIDS data live and how subjects/tasks are selected
- `design`: how first-level regressors are built
- `first_level`: what first-level outputs to write
- `pls`: how those outputs are mapped into a PLS analysis
- `execution`: local versus array execution settings
- `outputs`: where artifacts are written

## What is the smallest valid spec?

The minimum viable spec is deliberately small. You need:

- a BIDS directory
- at least one task label
- a first-level design formula
- a PLS method
- an output root

```{r minimal-spec, include = FALSE}
tmp <- tempfile("plsrri-yaml-")
dir.create(tmp, recursive = TRUE)
bids_dir <- file.path(tmp, "bids")
dir.create(bids_dir, recursive = TRUE)

minimal_spec <- list(
  dataset = list(
    bids_dir = bids_dir,
    task = "stroop"
  ),
  design = list(
    formula = "onset ~ hrf(condition, basis = 'spmg1')"
  ),
  first_level = list(
    output = list(
      type = "estimates",
      statistics = "estimate"
    )
  ),
  pls = list(
    method = "task",
    nperm = 0L,
    nboot = 0L
  ),
  outputs = list(
    root = file.path(tmp, "out")
  )
)

validated_minimal <- validate_pipeline_spec(minimal_spec)

stopifnot(
  identical(validated_minimal$pls$method, "task"),
  identical(validated_minimal$first_level$output$type, "estimates"),
  identical(validated_minimal$execution$mode, "local"),
  identical(validated_minimal$pls$input$statistic, "estimate")
)
```

```{r minimal-spec-print}
cat(as.yaml(minimal_spec))
```

That small object already validates and picks up defaults for `execution.mode`,
`execution.parallelism`, `first_level.strategy`, and `pls.input`.

## How should you read the top-level sections?

### `dataset`

`dataset` answers: what study is this, and which task/run space should be used?

Common fields:

- `bids_dir`
- `task`
- `space`
- `group_column`
- optional subject/session/run filters

```{r dataset-example, eval = FALSE}
dataset:
  bids_dir: /path/to/bids
  task: stroop
  space: MNI152NLin2009cAsym
  group_column: group
```

### `design`

`design` defines the first-level model, not the PLS contrast. The key field is
`formula`.

```{r design-example, eval = FALSE}
design:
  formula: onset ~ hrf(condition, basis = 'spmg1')
  block: ~ run
```

That formula is where you choose single-df HRFs versus basis expansions such as
FIR or tent-style models.

### `first_level`

`first_level` controls how the GLM stage is run and what maps are written.

```{r firstlevel-example, eval = FALSE}
first_level:
  strategy: runwise
  nchunks: 1
  output:
    type: estimates
    statistics: [estimate]
```

The important distinction is:

- `type = estimates`: write condition-level beta-like maps
- `type = contrasts`: write named contrast maps
- `type = F`: write F-statistic outputs

### `pls`

`pls` tells the second stage how to interpret first-level outputs.

```{r pls-example, eval = FALSE}
pls:
  method: task
  input:
    type: estimates
    statistic: estimate
  nperm: 1000
  nboot: 500
```

The `method` field maps onto supported `plsrri` methods such as:

- `task`
- `task_nonrotated`
- `behavior`
- `behavior_nonrotated`
- `multiblock`
- `multiblock_nonrotated`

### `execution`

`execution` controls how staged commands are scheduled, not the statistical
analysis itself.

```{r execution-example, eval = FALSE}
execution:
  mode: array
  parallelism: 8
```

### `outputs`

`outputs.root` is the artifact root that the CLI, scripted R API, report layer,
and Shiny attach mode all reuse.

```{r outputs-example, eval = FALSE}
outputs:
  root: plscli-out
```

## How do basis-expanded first-level outputs fit into the spec?

The YAML needs two pieces of information when first-level output labels encode
basis functions such as FIR bins or tent functions:

1. how first-level labels are written
2. how PLS should fold those labels back into a basis-aware manifest

```{r basis-spec, include = FALSE}
basis_spec <- list(
  dataset = list(
    bids_dir = bids_dir,
    task = "stroop"
  ),
  design = list(
    formula = "onset ~ hrf(condition, basis = 'fir', K = 4)"
  ),
  first_level = list(
    output = list(
      type = "estimates",
      statistics = "estimate",
      basis_pattern = "^(.*)_bin([0-9]+)$",
      basis_order = c("bin1", "bin2", "bin3", "bin4")
    )
  ),
  pls = list(
    method = "task",
    input = list(
      type = "estimates",
      statistic = "estimate",
      basis_pattern = "^(.*)_bin([0-9]+)$",
      condition_group = 1L,
      basis_group = 2L,
      basis_order = c("1", "2", "3", "4")
    ),
    nperm = 0L,
    nboot = 0L
  ),
  outputs = list(
    root = file.path(tmp, "fir-out")
  )
)

validated_basis <- validate_pipeline_spec(basis_spec)

stopifnot(
  identical(validated_basis$pls$input$condition_group, 1L),
  identical(validated_basis$pls$input$basis_group, 2L),
  length(validated_basis$pls$input$basis_order) == 4L
)
```

```{r basis-spec-print}
cat(as.yaml(basis_spec))
```

The important point is that basis handling belongs in both places:

- `design.formula` determines what the first-level model estimates
- `pls.input.*` tells the PLS stage how to reinterpret those basis-labelled maps

## How does the spec map to CLI stages?

The YAML is consumed incrementally. Not every stage needs every section.

| CLI stage | Main sections used |
|---|---|
| `plscli validate` | all |
| `plscli discover` | `dataset`, `outputs` |
| `plscli firstlevel-plan` | `dataset`, `design`, `first_level`, `outputs` |
| `plscli firstlevel-run` | first-level plan artifacts plus `execution` |
| `plscli pls-plan` | `pls`, `outputs` |
| `plscli pls-run` | `pls`, planned manifests, `outputs` |
| `plscli report` | `outputs` or an existing artifact root |

That split is why the same YAML can drive:

- a local end-to-end run
- an HPC array workflow
- a Shiny-exported pipeline configuration

## What should you treat as stable?

The stable contract is:

- the YAML structure described here
- the artifact root under `outputs.root`
- the public helpers:
  - `write_pipeline_template()`
  - `read_pipeline_spec()`
  - `validate_pipeline_spec()`
  - `prepare_firstlevel()`
  - `prepare_pls()`
  - `run_pls()`
  - `render_pls_report()`

## Where should you go next?

Use `vignette("scripted-workflows")` for the staged R and CLI workflow, and
`?write_pipeline_template` / `?read_pipeline_spec` for the corresponding help
pages.
