---
title: "Getting Started with plsrri"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with plsrri}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)

ensure_local_plsrri <- function(required_exports) {
  pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."
  if (!file.exists(file.path(pkg_root, "DESCRIPTION")) ||
      !requireNamespace("devtools", quietly = TRUE)) {
    return(invisible(FALSE))
  }

  installed_exports <- if (requireNamespace("plsrri", quietly = TRUE)) {
    getNamespaceExports("plsrri")
  } else {
    character(0)
  }

  if (!all(required_exports %in% installed_exports)) {
    devtools::load_all(pkg_root, quiet = TRUE)
    return(invisible(TRUE))
  }

  invisible(FALSE)
}

ensure_local_plsrri("quick_pls")
```

```{r load-package}
library(plsrri)
```

> **Note:** `plsrri` is under active development. The API may change before the first stable release.

You have subject-by-condition brain data and want to know two things: which
latent patterns separate your experimental conditions, and how stable are
those patterns under resampling? `plsrri` gives you both a direct entry
point for standard analyses and a builder workflow when you need more
control.

## How does a PLS analysis work?

The core workflow has three steps:

1. **Specify** — describe your data, conditions, groups, and analysis
   parameters in a `pls_spec` object
2. **Run** — execute the analysis to get a `pls_result` containing singular
   values, saliences, scores, and optional inference (permutation, bootstrap)
3. **Extract** — pull out what you need with accessor functions:
   `significance()`, `salience()`, `bsr()`, `scores()`, `loadings()`,
   `confidence()`

| Object | What it holds | How you get it |
|:-------|:-------------|:---------------|
| `pls_spec` | Data matrices, method, resampling config | `pls_spec()` + builder verbs |
| `pls_result` | SVD decomposition, scores, inference | `run()` or `quick_pls()` |

The fastest path is `quick_pls()`, which wraps all three steps into one call.
The builder API (`pls_spec() |> add_subjects() |> configure() |> run()`)
gives you labels, metadata, and full control over every parameter.

## What does a minimal analysis look like?

This vignette uses a small synthetic dataset: two groups (young, older),
three conditions, and 200 voxels with planted signal in two blocks.

```{r simulate-task-data, include = FALSE}
set.seed(123)

n_subj <- 15
n_cond <- 3
n_vox <- 200

make_group <- function(effect_size) {
  signal <- matrix(0, nrow = n_subj * n_cond, ncol = n_vox)

  for (s in seq_len(n_subj)) {
    row_enc <- (s - 1L) * n_cond + 1L
    row_ret <- (s - 1L) * n_cond + 2L
    signal[row_enc, 1:80] <- effect_size
    signal[row_ret, 81:140] <- effect_size * 0.7
  }

  signal + matrix(rnorm(n_subj * n_cond * n_vox), ncol = n_vox)
}

group_young <- make_group(effect_size = 1.0)
group_older <- make_group(effect_size = 0.5)
```

Each group matrix has `r n_subj` subjects $\times$ `r n_cond` conditions =
`r n_subj * n_cond` rows and `r n_vox` voxels:

```{r inspect-task-data}
rbind(
  young = dim(group_young),
  older = dim(group_older)
)
```

`quick_pls()` is the shortest path. It runs mean-centering Task PLS with
permutation and bootstrap inference in one call:

```{r run-task-pls}
result <- quick_pls(
  datamat_lst = list(group_young, group_older),
  num_subj_lst = c(n_subj, n_subj),
  num_cond = n_cond,
  nperm = 200,
  nboot = 200,
  progress = FALSE
)
```

The result object carries everything: `r n_lv(result)` latent variables,
each with a p-value from the permutation test and voxel-level bootstrap
ratios.

## Which latent variable matters?

Start with `significance()` to see which latent variables survive the
permutation test, and `singular_values()` to see how variance is distributed:

```{r task-significance}
cbind(
  pvalue = round(significance(result), 3),
  variance = round(singular_values(result, normalize = TRUE), 1)
)
```

LV1 captures `r round(singular_values(result, normalize = TRUE)[1], 1)`%
of the variance. The scree plot makes this dominance visual:

```{r plot-task-singular-values, fig.cap = "Variance explained by each latent variable. LV1 dominates, consistent with the planted signal."}
plot_singular_values(result)
```

## How do conditions load onto LV1?

Design scores show how each group-by-condition cell projects onto the latent
variable. Large positive and negative bars indicate conditions that drive the
pattern in opposite directions.

```{r plot-task-design-scores, fig.cap = "Design scores for LV1. Encoding (condition 1) loads positively; retrieval (condition 2) loads in the opposite direction."}
plot_scores(result, lv = 1, type = "design", plot_type = "bar")
```

The separation between conditions confirms that LV1 captures a genuine
experimental effect, not noise.

## How consistent are subjects?

Brain scores show whether individual subjects express the same latent
pattern. Tight distributions mean the effect is reliable; wide spread
suggests individual differences.

```{r plot-task-brain-scores, fig.cap = "Brain scores for LV1. Each point is one subject-condition observation."}
plot_scores(result, lv = 1, type = "brain", plot_type = "violin")
```

## Which voxels contribute reliably?

Bootstrap ratios (BSRs) are saliences divided by their bootstrap standard
errors — they work like z-scores. Voxels with $|BSR| > 2$ or $3$ are
typically considered reliable.

```{r task-bsr}
lv1_bsr <- bsr(result, lv = 1)
reliable <- sum(abs(lv1_bsr) > 2)
```

`r reliable` of `r n_features(result)` voxels have $|BSR| > 2$. A quick
profile shows where the reliable voxels cluster:

```{r plot-task-bsr, echo = FALSE, fig.cap = "Bootstrap ratio profile across voxels. The horizontal lines mark the |BSR| = 2 threshold. Voxels 1-80 and 81-140 show the planted signal blocks.", fig.height = 3}
plot(as.numeric(lv1_bsr), type = "h", col = "steelblue",
     xlab = "Voxel", ylab = "Bootstrap Ratio (LV1)",
     main = NULL, las = 1)
abline(h = c(-2, 2), lty = 2, col = "grey50")
```

You can also threshold directly to get a cleaned vector:

```{r task-bsr-threshold}
lv1_bsr_clean <- bsr(result, lv = 1, threshold = 2)
sum(lv1_bsr_clean != 0)
```

## When should you use the builder workflow?

Use the builder API when you want condition labels, group names, or
non-default method settings to travel with the analysis. The pipe chain
makes the full specification explicit and readable:

```{r run-builder-workflow}
builder_result <- pls_spec() |>
  add_subjects(
    list(group_young, group_older),
    groups = c(n_subj, n_subj)
  ) |>
  add_conditions(n_cond, labels = c("encoding", "retrieval", "rest")) |>
  add_group_labels(c("young", "older")) |>
  configure(method = "task", nperm = 200, nboot = 200) |>
  run(progress = FALSE)
```

The numerical result is identical to `quick_pls()`, but now plots carry
meaningful labels:

```{r plot-builder-scores, fig.cap = "Design scores with condition and group labels from the builder workflow."}
plot_scores(builder_result, lv = 1, type = "design", plot_type = "bar")
```

The builder API also supports behavior data (`add_behavior()`), design
contrasts (`add_design()`), and non-rotated methods — see
`vignette("behavior-pls")` for the behavior workflow.

## What about the interactive GUI?

For exploratory work, `plsrri` ships a Shiny application that wraps the
full setup-analyze-explore workflow in a point-and-click interface:
```r
launch_pls_gui()
```
The GUI supports loading data from BIDS directories, attaching to
CLI pipeline outputs, and interactive brain visualization with volume and
surface views.

## Where to go next

If you want to move from interactive examples to reproducible scripted analysis, go next to `vignette("scripted-workflows")`, which shows the public non-GUI contract built around saved first-level artifacts, staged CLI runs, and Quarto reporting.

| Goal | Resource |
|:-----|:---------|
| Scripted R API, CLI, and report workflow | `vignette("scripted-workflows")` |
| Behavior PLS with continuous measures | `vignette("behavior-pls")` |
| Multisite pooled behavior PLS with site diagnostics | `vignette("site-pooling")` |
| Within-subject seed connectivity from trial-level data | `vignette("ws-seed-pls")` |
| Multiblock and seed connectivity PLS | `vignette("multiblock-and-seed")` |
| Full engine parameters | `?pls_analysis` |
| Builder verbs | `?pls_spec`, `?configure`, `?run` |
| Plotting functions | `?plot_scores`, `?plot_loadings`, `?plot_singular_values` |
| Bootstrap and permutation details | `?bsr`, `?significance`, `?confidence` |
