---
title: "Predictive Measures and Resampling Inference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Predictive Measures and Resampling Inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)

ensure_local_plsrri <- function(required_exports) {
  pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."
  if (!file.exists(file.path(pkg_root, "DESCRIPTION")) ||
      !requireNamespace("devtools", quietly = TRUE)) {
    return(invisible(FALSE))
  }

  installed_exports <- if (requireNamespace("plsrri", quietly = TRUE)) {
    getNamespaceExports("plsrri")
  } else {
    character(0)
  }

  if (!all(required_exports %in% installed_exports)) {
    devtools::load_all(pkg_root, quiet = TRUE)
    return(invisible(TRUE))
  }

  invisible(FALSE)
}

ensure_local_plsrri("evaluate_prediction")

round_numeric_df <- function(df, digits = 2) {
  num <- vapply(df, is.numeric, logical(1))
  df[num] <- lapply(df[num], round, digits = digits)
  df
}
```

```{r load-package}
library(plsrri)
```

You can now ask two different questions of the same PLS analysis. First, does
the latent-variable decomposition look non-random and stable under the usual
PLS resampling tools? Second, can the fitted score space support honest
out-of-fold prediction for held-out subjects? `plsrri` now supports both.

This article shows how predictive measures complement, rather than replace,
the existing permutation, bootstrap, and split-half analyses:

- permutation on a `pls_result` asks whether an LV is stronger than expected
  under the null decomposition
- bootstrap asks which voxel or behavior relationships are stable
- predictive cross-validation asks whether the learned score space
  generalizes to held-out subjects

## What does a predictive quick win look like?

This vignette uses one synthetic brain-behavior dataset throughout. The quick
win is a classification analysis: can you recover a held-out high-vs-low
subject label from the score space learned inside each training fold?

```{r simulate-and-fit, include = FALSE}
set.seed(2026)

n_subj <- 24L
n_cond <- 2L
n_feat <- 90L

subj_latent <- as.numeric(scale(
  seq(-1.5, 1.5, length.out = n_subj) + rnorm(n_subj, sd = 0.2)
))
cond_shift <- c(-0.5, 0.5)

behav <- cbind(
  symptom = as.numeric(scale(
    rep(subj_latent, times = n_cond) +
      rep(cond_shift, each = n_subj) +
      rnorm(n_subj * n_cond, sd = 0.15)
  )),
  reserve = as.numeric(scale(
    -0.7 * rep(subj_latent, times = n_cond) +
      rnorm(n_subj * n_cond, sd = 0.2)
  ))
)

brain <- matrix(
  rnorm(n_subj * n_cond * n_feat, sd = 0.5),
  nrow = n_subj * n_cond,
  ncol = n_feat
)

brain[, 1:30] <- brain[, 1:30] +
  rep(subj_latent, times = n_cond) %o% rep(1.0, 30)
brain[, 31:60] <- brain[, 31:60] +
  rep(cond_shift, each = n_subj) %o% rep(0.8, 30)
brain[, 61:90] <- brain[, 61:90] +
  rep(subj_latent * rep(cond_shift, each = n_subj), 1) %o% rep(0.6, 30)

symptom_score <- subj_latent + rnorm(n_subj, sd = 0.05)
diagnosis <- factor(
  ifelse(subj_latent > median(subj_latent), "high", "low"),
  levels = c("low", "high")
)

spec <- pls_spec() |>
  add_subjects(list(brain), groups = n_subj) |>
  add_conditions(c("baseline", "challenge")) |>
  add_behavior(behav, measures = c("symptom", "reserve")) |>
  configure(method = "behavior", nperm = 79, nboot = 79)

fit <- run(spec, progress = FALSE)

reg_result <- evaluate_prediction(
  spec,
  outcome = symptom_score,
  task = "regression",
  components = 1:2,
  resampling = list(
    outer_folds = 4L,
    outer_repeats = 1L,
    inner_folds = 3L,
    inner_repeats = 1L,
    stratify = FALSE
  ),
  num_perm = 9L,
  num_boot = 29L,
  seed = 11L,
  progress = FALSE
)

cls_result <- evaluate_prediction(
  spec,
  outcome = diagnosis,
  task = "classification",
  components = 1:2,
  resampling = list(
    outer_folds = 4L,
    outer_repeats = 1L,
    inner_folds = 3L,
    inner_repeats = 1L,
    stratify = TRUE
  ),
  num_perm = 9L,
  num_boot = 29L,
  seed = 12L,
  progress = FALSE
)

sig <- significance(fit)
var_exp <- singular_values(fit, normalize = TRUE)

corr_ci <- confidence(fit, what = "correlation", lv = 1)
corr_labels <- as.vector(sapply(spec$conditions, function(cond) {
  paste(cond, colnames(behav), sep = " / ")
}))
corr_table <- data.frame(
  loading = corr_labels,
  lower = corr_ci$lower[, 1],
  upper = corr_ci$upper[, 1],
  stringsAsFactors = FALSE
)
reliable_corr <- sum(corr_table$lower * corr_table$upper > 0)

reg_summary <- setNames(reg_result$summary$mean, reg_result$summary$metric)
cls_summary <- setNames(cls_result$summary$mean, cls_result$summary$metric)

reg_subject <- aggregate(
  cbind(estimate, truth) ~ subject_id,
  data = reg_result$oof_predictions,
  FUN = mean
)

cls_subject <- aggregate(
  probability ~ subject_id + truth,
  data = cls_result$oof_predictions,
  FUN = mean
)

comparison <- data.frame(
  question = c(
    "Is the dominant latent variable stronger than a null decomposition?",
    "Which behavior relationships are stable under bootstrap resampling?",
    "Can the latent score space predict a held-out continuous outcome?",
    "Can the latent score space classify held-out subjects?"
  ),
  diagnostic = c(
    sprintf("LV1 permutation p = %.3f; variance explained = %.1f%%", sig[1], var_exp[1]),
    sprintf("%d of %d behavior intervals exclude zero", reliable_corr, nrow(corr_table)),
    sprintf("OOF R^2 = %.2f; RMSE = %.2f", reg_summary[["rsq"]], reg_summary[["rmse"]]),
    sprintf("OOF accuracy = %.2f; AUC = %.2f", cls_summary[["accuracy"]], cls_summary[["auc"]])
  ),
  stringsAsFactors = FALSE
)

stopifnot(
  all(is.finite(sig)),
  sig[1] < 0.05,
  is.finite(var_exp[1]),
  var_exp[1] > 50,
  reliable_corr >= 4,
  all(is.finite(reg_result$summary$mean)),
  reg_summary[["rsq"]] > 0.5,
  reg_result$perm_test$observed < median(reg_result$perm_test$null_distribution),
  all(reg_result$boot_ci$lower <= reg_result$boot_ci$upper),
  all(is.finite(cls_result$summary$mean)),
  cls_summary[["accuracy"]] > 0.7,
  cls_summary[["auc"]] > 0.75,
  cls_result$perm_test$observed > median(cls_result$perm_test$null_distribution),
  all(cls_result$boot_ci$lower <= cls_result$boot_ci$upper),
  nrow(reg_subject) == n_subj,
  nrow(cls_subject) == n_subj
)
```

```{r quick-win}
round_numeric_df(cls_result$summary)
```

```{r quick-win-inference}
round_numeric_df(cls_result$boot_ci[, c("metric", "estimate", "lower", "upper")])
```

The key point is that this object is not another decomposition summary. It is
an out-of-fold predictive result built from nested cross-validation, with its
own permutation null and bootstrap intervals.

## What data are we using?

The example has `r n_subj` subjects, `r n_cond` conditions, `r n_feat`
features, and two behavior measures. One subject-level latent factor drives a
brain block, the symptom measure, and the binary diagnosis; a second block
captures condition structure.

```{r inspect-data}
rbind(
  brain = dim(brain),
  behavior = dim(behav)
)
```

```{r inspect-subject-outcomes}
head(data.frame(
  subject = seq_len(6),
  symptom_score = round(symptom_score[1:6], 2),
  diagnosis = diagnosis[1:6]
))
```

## What does ordinary PLS inference tell you first?

Start with the decomposition-side questions. Here the behavior PLS fit asks
whether there is a dominant brain-behavior latent variable, and whether the
behavior links are stable under bootstrap resampling.

```{r decomposition-significance}
round(cbind(
  pvalue = significance(fit),
  variance = singular_values(fit, normalize = TRUE)
), 2)
```

```{r plot-scree, fig.cap = "The decomposition-side permutation test is about latent variables, not held-out prediction. LV1 dominates the singular value spectrum."}
plot_singular_values(fit)
```

LV1 is clearly non-null in the ordinary PLS sense. That still does not tell
you whether held-out subjects can be predicted. It only tells you that the
brain-behavior decomposition is stronger than the null decomposition generated
by row shuffling.

Bootstrap intervals answer a different question again: which specific
brain-behavior links are stable?

```{r behavior-correlation-ci}
round_numeric_df(corr_table)
```

All four behavior-by-condition intervals stay away from zero. That supports
the claim that the symptom/reserve pattern is stable, but it is still not a
held-out predictive statement.

## What does predictive regression add?

Now switch from decomposition inference to honest prediction. The regression
target is the subject-level symptom score, not the repeated behavior matrix
used to fit the decomposition.

```{r regression-summary}
round_numeric_df(reg_result$summary)
```

```{r regression-inference}
round_numeric_df(reg_result$boot_ci[, c("metric", "estimate", "lower", "upper")])
```

For speed, this vignette uses only `r reg_result$perm_test$num_perm`
predictive label permutations, so the smallest attainable predictive
p-value is `r 1 / (reg_result$perm_test$num_perm + 1)`. In a real analysis
you would increase this substantially.

```{r regression-permutation}
reg_result$perm_test
```

The continuous outcome is recoverable out of sample: the held-out
`R^2` is `r sprintf('%.2f', reg_summary[['rsq']])`, and the observed RMSE is
smaller than the median value from the shuffled-label null.

```{r plot-regression-oof, echo = FALSE, fig.cap = "Each point is one held-out subject, aggregated across outer folds. Predictive validation asks whether these subject-level estimates track the true outcome."}
plot(
  reg_subject$truth,
  reg_subject$estimate,
  pch = 19,
  col = "steelblue",
  xlab = "True symptom score",
  ylab = "Out-of-fold predicted symptom score"
)
abline(a = 0, b = 1, lty = 2, col = "grey40")
```

## What does predictive classification add?

The same score space can also be used for classification. Here the target is
a binary high-vs-low diagnosis derived from the same latent factor.

```{r classification-summary}
round_numeric_df(cls_result$summary)
```

```{r classification-permutation}
cls_result$perm_test
```

```{r classification-boot}
round_numeric_df(cls_result$boot_ci[, c("metric", "estimate", "lower", "upper")])
```

Again, the predictive inference is about held-out performance. Accuracy and
AUC are computed from out-of-fold predictions, and the predictive permutation
test compares those numbers to a null generated by re-running the full nested
cross-validation workflow with shuffled subject labels.

```{r plot-classification-oof, echo = FALSE, fig.cap = "Held-out class probabilities aggregated to one row per subject. The horizontal line marks a 0.5 decision threshold."}
boxplot(
  probability ~ truth,
  data = cls_subject,
  col = c("grey85", "tomato"),
  ylab = "Mean held-out Pr(high)",
  xlab = "True class"
)
abline(h = 0.5, lty = 2, col = "grey40")
```

## How do the inferential layers complement each other?

These tools are strongest when you treat them as complementary rather than
interchangeable:

```{r comparison-table}
comparison
```

You can read the table as four distinct questions:

- Decomposition permutation: is there a latent variable stronger than a null
  PLS decomposition?
- Decomposition bootstrap: which loadings or correlations are stable?
- Predictive regression/classification: do those learned score patterns
  generalize to held-out subjects?

Split-half validation fits beside the first two, not the last one. It asks
whether the decomposition itself replicates across random halves of the data,
while `evaluate_prediction()` asks whether a training-fit score space supports
honest out-of-fold prediction.

## Where should you go next?

| Goal | Resource |
|:-----|:---------|
| Core task PLS workflow | `vignette("plsrri")` |
| Ordinary behavior PLS workflow | `vignette("behavior-pls")` |
| Multiblock and seed methods | `vignette("multiblock-and-seed")` |
| Predictive entry point | `?evaluate_prediction` |
| Decomposition-side significance | `?significance` |
| Bootstrap confidence intervals | `?confidence` |
