---
title: "Scripted End-to-End Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scripted End-to-End Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)

needed <- c("RNifti", "neuroim2", "yaml", "quarto")
missing_pkgs <- needed[!vapply(needed, requireNamespace, quietly = TRUE, FUN.VALUE = logical(1))]
if (length(missing_pkgs) > 0) {
  knitr::knit_exit()
}

pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."
if (requireNamespace("plsrri", quietly = TRUE) &&
    !"prepare_pls" %in% getNamespaceExports("plsrri") &&
    requireNamespace("devtools", quietly = TRUE) &&
    file.exists(file.path(pkg_root, "DESCRIPTION"))) {
  devtools::load_all(pkg_root, quiet = TRUE)
}
```

```{r load-package}
library(plsrri)
```

You want one analysis contract that works in three places: ordinary R code, a staged CLI workflow, and later the Shiny app. In `plsrri`, that contract is a saved artifact directory containing first-level manifests, a planned PLS manifest, a saved `pls_result`, and optional reports.

This vignette starts from a tiny synthetic artifact root so the core workflow stays runnable under `R CMD check`. It then shows how the same workflow extends to `prepare_firstlevel()` and `plscli` on a real BIDS dataset.

## What does the saved-artifact workflow look like?

The smallest stable path is:

1. point `prepare_pls()` at a saved artifact root or pipeline YAML
2. run the returned `pls_spec`
3. render a Quarto report from the same saved artifacts

For performance-sensitive scripted runs, `plsrri` also exposes exact fast-path
controls through the global option `plsrri.fast_paths`. The practical default is
to enable the well-performing exact accelerations and leave the slower
permutation branch off:

```{r fast-path-options, eval = FALSE}
options(plsrri.fast_paths = c("xcor", "bootstrap"))

# Opt into the exact permutation fast path as well.
options(plsrri.fast_paths = c("xcor", "bootstrap", "permutation"))

# Disable exact fast paths entirely.
options(plsrri.fast_paths = FALSE)
```

That split is deliberate: current benchmarks show exact reduced-space bootstrap
is a substantial win, while the exact permutation fast path is still best treated
as opt-in.

```{r build-synthetic-artifacts, include = FALSE}
set.seed(2026)

tmp <- tempfile("plsrri-scripted-")
dir.create(tmp, recursive = TRUE)
bids_dir <- file.path(tmp, "bids")
out_root <- file.path(tmp, "out")
dir.create(bids_dir, recursive = TRUE)

mask_arr <- array(0, dim = c(2, 2, 2))
mask_arr[1, 1, 1] <- 1
mask_arr[2, 1, 1] <- 1
mask_file <- file.path(tmp, "mask.nii.gz")
RNifti::writeNifti(mask_arr, mask_file)

subjects <- c("01", "02")
scans <- file.path(bids_dir, paste0("sub-", subjects, "_task-toy_bold.nii.gz"))
events <- file.path(bids_dir, paste0("sub-", subjects, "_task-toy_events.tsv"))
invisible(file.create(scans))
for (event_file in events) {
  writeLines(c("onset\tduration\tcondition", "0\t1\tface", "2\t1\tscene"), event_file)
}

dir.create(file.path(out_root, "firstlevel", "work"), recursive = TRUE)

work_ids <- sprintf("w%04d", seq_along(subjects))
work_plan <- data.frame(
  subject = subjects,
  group = "all",
  task = "toy",
  work_id = work_ids,
  scan_count = 1L,
  work_dir = file.path("firstlevel", "work", work_ids),
  fit_file = file.path("firstlevel", "work", work_ids, "fit.rds"),
  manifest_file = file.path("firstlevel", "work", work_ids, "maps.tsv"),
  stringsAsFactors = FALSE
)
utils::write.table(
  work_plan,
  file.path(out_root, "firstlevel", "work_plan.tsv"),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)

values <- list(
  "01" = c(face = 2.4, scene = -2.3),
  "02" = c(face = 2.0, scene = -1.9)
)

for (i in seq_along(subjects)) {
  work_dir_abs <- file.path(out_root, "firstlevel", "work", work_ids[[i]])
  dir.create(work_dir_abs, recursive = TRUE, showWarnings = FALSE)

  manifest_rows <- lapply(names(values[[subjects[[i]]]]), function(label) {
    arr <- array(0, dim = c(2, 2, 2))
    arr[1, 1, 1] <- values[[subjects[[i]]]][[label]]
    arr[2, 1, 1] <- values[[subjects[[i]]]][[label]] + 0.5
    out_file <- file.path(work_dir_abs, paste0(label, ".nii.gz"))
    RNifti::writeNifti(arr, out_file)

    data.frame(
      work_id = work_ids[[i]],
      subject = subjects[[i]],
      group = "all",
      task = "toy",
      type = "estimates",
      label = label,
      statistic = "estimate",
      file = file.path("firstlevel", "work", work_ids[[i]], paste0(label, ".nii.gz")),
      mask_file = mask_file,
      condition = label,
      basis_label = NA_character_,
      basis_index = NA_integer_,
      stringsAsFactors = FALSE
    )
  })

  utils::write.table(
    do.call(rbind, manifest_rows),
    file.path(work_dir_abs, "maps.tsv"),
    sep = "\t",
    quote = FALSE,
    row.names = FALSE
  )
}

spec_path <- file.path(tmp, "pipeline.yml")
yaml::write_yaml(
  list(
    dataset = list(
      bids_dir = bids_dir,
      task = "toy",
      mask = mask_file
    ),
    design = list(
      formula = "onset ~ hrf(condition, basis = 'spmg1')"
    ),
    first_level = list(
      output = list(
        type = "estimates",
        statistics = "estimate"
      )
    ),
    pls = list(
      method = "task",
      nperm = 0L,
      nboot = 0L,
      nsplit = 0L
    ),
    outputs = list(
      root = out_root
    )
  ),
  spec_path
)
```

```{r prepare-pls-from-artifacts}
pspec <- prepare_pls(out_root, pls_options = list(method = "task", nperm = 0L, nboot = 0L))

stopifnot(
  inherits(pspec, "pls_spec"),
  pspec$num_cond == 2L,
  length(pspec$datamat_lst) == 1L
)

pspec
```

Once you have a prepared spec, `run_pls()` and `render_pls_report()` are thin, deterministic wrappers.

```{r run-and-report}
result <- run_pls(spec_path)
report_file <- file.path(tmp, "plsrri-scripted-report.html")
rendered <- render_pls_report(spec_path, output_file = report_file, include_brain = FALSE)

stopifnot(
  inherits(result, "pls_result"),
  all(is.finite(result$s)),
  file.exists(file.path(out_root, "pls", "pls_result.rds")),
  file.exists(rendered),
  file.info(rendered)$size > 0
)

c(
  latent_variables = n_lv(result),
  features = n_features(result)
)
```

The saved result and the rendered report are part of the same artifact contract, so you can reopen them later without reconstructing the analysis state.

```{r inspect-saved-artifacts}
summary_file <- file.path(out_root, "pls", "pls_summary.tsv")
saved_summary <- utils::read.delim(summary_file, sep = "\t", stringsAsFactors = FALSE)

stopifnot(
  nrow(saved_summary) >= 1L,
  all(is.finite(saved_summary$singular_value))
)

saved_summary
```

## When do you call `prepare_firstlevel()`?

Use `prepare_firstlevel()` when you want the same R session to own BIDS discovery, first-level planning, and first-level execution before moving into PLS.

That live path depends on the external first-level stack (`bidser`, `fmridataset`, `fmridesign`, `fmrireg`), so it is shown here as an execution pattern rather than as a vignette-evaluated chunk:

```r
prep <- prepare_firstlevel("study.yml")
result <- run_pls(prep)
render_pls_report(prep$spec)
```

For workstation use, that is often the cleanest end-to-end R workflow. For HPC use, the same YAML spec can be split into explicit stages.

## How does the same workflow look from the CLI?

The CLI mirrors the same contract and the same stage boundaries.

```bash
plscli validate --spec study.yml
plscli discover --spec study.yml
plscli firstlevel-plan --spec study.yml
plscli firstlevel-run --spec study.yml --work-id w0007
plscli pls-plan --spec study.yml
plscli pls-run --spec study.yml
plscli report --input study.yml --format html
```

For array jobs, the only thing that changes is how you schedule `firstlevel-run`.

```bash
plscli firstlevel-plan --spec study.yml
plscli firstlevel-run --spec study.yml --work-id "${SLURM_ARRAY_TASK_ID}"
plscli pls-plan --spec study.yml
plscli pls-run --spec study.yml
plscli report --input /path/to/artifact-root --format pdf
```

The important design point is that `plscli report` reads the same artifacts that `render_pls_report()` reads in R, and `prepare_pls()` reads the same first-level manifests that the CLI writes.

## What should you treat as the public contract?

For scripted analysis, the stable surfaces are:

- `prepare_firstlevel()`
- `prepare_pls()`
- `run_pls()`
- `render_pls_report()`

The stable on-disk contract is the artifact root containing:

- `firstlevel/work_plan.tsv`
- `firstlevel/work/*/maps.tsv`
- `pls/pls_manifest.tsv`
- `pls/pls_result.rds`
- `pls/pls_summary.tsv`
- `reports/`

That is the contract the Shiny app is also expected to consume when it attaches to existing first-level outputs.

## Where should you go next?

Use `vignette("plsrri")` for the main task-PLS workflow, `vignette("behavior-pls")` for behavior PLS, and `vignette("multiblock-and-seed")` for multiblock and seed analyses.

The most relevant help pages after this vignette are:

- `vignette("pipeline-yaml-spec")`
- `?prepare_firstlevel`
- `?prepare_pls`
- `?run_pls`
- `?render_pls_report`
- `?plscli_main`
