---
title: "Mean-Centered Task PLS for First-Level SDAM Maps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mean-Centered Task PLS for First-Level SDAM Maps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)

run_full <- identical(Sys.getenv("PLSRRI_RUN_SDAM_VIGNETTE"), "true")
knitr::opts_chunk$set(eval = run_full)
```

This example starts from first-level image maps that are already organized by
subject, task, and measure. The goal is to test whether a mean-centered Task
PLS can find reliable multivariate patterns across two tasks (`recog`,
`nback`), two map measures (`low_mid`, `high_sem`), and two participant groups
(`control`, `sdam`).

The maps are pre-masked. Instead of pretending that zeros are valid signal, we
derive a common finite nonzero mask across all maps and use that mask for image
ingestion, bootstrap-ratio maps, and plotting.

## What do the inputs look like?

The complete runnable code is installed with the package as an example script.
It defines small helpers for the manifest, mask, model specification, and
figures.

```{r load-example}
library(plsrri)

source(system.file(
  "examples",
  "sdam_firstlevel_task_pls.R",
  package = "plsrri"
))
```

Build one row per subject-condition map. Here, "condition" means the
combination of task and measure.

```{r build-manifest}
root <- sdam_testdata_root("testdata")
manifest <- build_sdam_manifest(root)

summarise_sdam_design(manifest)
```

The manifest has 128 image rows: 32 subjects, four task-by-measure conditions,
and two groups. The order is stored explicitly with factor levels so that PLS
scores and loadings use stable labels.

## How do we handle the missing mask?

The files are already pre-masked, so zero-valued voxels outside the usable data
region should not enter the analysis. The helper below keeps voxels that are
finite and nonzero in every map.

```{r derive-mask}
mask <- derive_common_nonzero_mask(manifest)

sum(as.array(mask) > 0)
```

Using the common mask gives a conservative feature space shared by all
subjects, tasks, and measures.

## How do we specify the PLS?

The model is a mean-centered Task PLS. Rows are grouped by participant group,
and each group has four conditions:

```{r configure-pls}
spec <- make_sdam_task_pls_spec(
  manifest = manifest,
  mask = mask,
  nperm = 1000,
  nboot = 500,
  meancentering = "within_group"
)

spec
```

`nperm` controls component-level inference for latent variables. `nboot`
controls voxel-level reliability estimates through bootstrap ratios.

## How do we run the analysis?

Set the random seed before fitting so the permutation and bootstrap samples are
reproducible.

```{r fit-pls}
set.seed(SDAM_SEED)
result <- run(spec, progress = TRUE)
```

Start with the latent-variable summary. A latent variable is worth
interpreting when it combines meaningful variance with a small permutation
p-value.

```{r summarize-result}
lv_summary <- summarise_sdam_result(result)
lv_summary
```

```{r choose-lv, include = FALSE}
lv_to_show <- if (any(lv_summary$significant)) {
  lv_summary$lv[which(lv_summary$significant)[[1]]]
} else {
  1L
}
condition_key <- sdam_condition_key()
```

## Which latent variables are reliable?

The singular-value plot shows variance explained and highlights permutation
significance.

```{r plot-singular-values, fig.cap = "Variance explained by latent variable, with permutation significance from the mean-centered Task PLS."}
plot_singular_values(result)
```

For an interpretable latent variable, inspect the design scores first. They
show which group-by-condition cells sit at opposite ends of the multivariate
pattern.

```{r plot-design-scores, fig.cap = "Design scores for the selected reliable LV across group-by-condition cells."}
plot_scores(result, lv = lv_to_show, type = "design", plot_type = "bar")
```

When the design has several crossed factors, the flattened bar plot is only
the starting point. Add a condition key to restore the task and level factors,
then view the same LV scores as a small factorial table.

```{r plot-design-heatmap, fig.cap = "The same design scores rehydrated as task-by-level heatmaps within group."}
plot_design_heatmap(
  result,
  lv = lv_to_show,
  condition_key = condition_key,
  row = "level",
  column = "task",
  facet = "group"
)
```

The interaction view is often easier to read when a latent variable mixes
group, task, and level effects.

```{r plot-design-interaction, fig.cap = "Design-score interaction plot for the selected LV."}
plot_design_interaction(
  result,
  lv = lv_to_show,
  condition_key = condition_key,
  x_axis = "task",
  trace = "level",
  facet = "group"
)
```

Finally, decompose the design vector into effect-coded factorial terms. The
largest bars show which main effects or interactions dominate the LV design
pattern. The LV sign is arbitrary, so use signs to connect the score plot to
the brain map, and use magnitudes to rank the terms.

```{r plot-design-contrasts, fig.cap = "Effect-coded decomposition of the selected LV design scores."}
plot_design_contrasts(result, lv = lv_to_show, condition_key = condition_key)
```

Brain scores show how consistently individual subject-condition observations
express the same latent pattern.

```{r plot-brain-scores, fig.cap = "Brain scores for the selected LV. Each point is one subject-condition observation."}
plot_scores(result, lv = lv_to_show, type = "brain", plot_type = "violin")
```

## Where is the reliable brain pattern?

Bootstrap ratios divide saliences by their bootstrap standard errors. The
first thresholded map keeps voxels with large absolute bootstrap ratios in the
original volume space.

```{r plot-bsr-volume, fig.cap = "Bootstrap-ratio map for the selected LV thresholded at |BSR| > 3 in volume space."}
plot_sdam_bsr_volume(result, lv = lv_to_show, threshold = 3)
```

For a surface-oriented view, the same thresholded bootstrap-ratio map can be
projected onto an fsaverage6 Schaefer surface atlas with
`neuroatlas::plot_brain()`. This is useful for visual comparison, but it is a
projection of the voxel map, not a different statistical test. The helper uses
a neutral grey base surface so the colours encode only the projected
bootstrap-ratio overlay.

```{r plot-bsr-surface, fig.cap = "Bootstrap-ratio map for the selected LV projected to an fsaverage6 surface with neuroatlas."}
plot_sdam_bsr_surface(result, lv = lv_to_show, threshold = 3)
```

You can save the fitted object, a latent-variable summary, and the figures with
the standalone script:

```sh
Rscript inst/examples/sdam_firstlevel_task_pls.R
```

By default the script writes to `artifacts/sdam-firstlevel-task-pls/`. Use
`PLSRRI_SDAM_TESTDATA`, `PLSRRI_SDAM_NPERM`, and `PLSRRI_SDAM_NBOOT` to point at
a different data directory or adjust inference counts.
