---
title: "Multisite Behavioral PLS Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multisite Behavioral PLS Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)

ensure_local_plsrri <- function(required_exports) {
  pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."
  if (!file.exists(file.path(pkg_root, "DESCRIPTION")) ||
      !requireNamespace("devtools", quietly = TRUE)) {
    return(invisible(FALSE))
  }

  installed_exports <- if (requireNamespace("plsrri", quietly = TRUE)) {
    getNamespaceExports("plsrri")
  } else {
    character(0)
  }

  if (!all(required_exports %in% installed_exports)) {
    devtools::load_all(pkg_root, quiet = TRUE)
    return(invisible(TRUE))
  }

  invisible(FALSE)
}

ensure_local_plsrri(c("add_site_labels", "site_pooling_diagnostics"))
```

```{r load-package}
library(plsrri)
```

If you pool subjects from multiple sites, a single behavioral PLS fit is often
the right starting point, but only if you can also ask whether the latent
pattern is stable across sites. This workflow shows a basic multisite strategy:

1. standardize brain and behavior within site
2. fit one pooled behavioral PLS model
3. inspect site-specific stability diagnostics after the fit

## What does a pooled multisite dataset look like?

For a pooled analysis you still need the standard behavior-PLS row structure:
subjects within conditions. The only addition is a site label for each subject
or observation.

```{r simulate-site-data, include = FALSE}
set.seed(20260401)

n_sites <- 3L
n_site_subj <- 12L
n_subj <- n_sites * n_site_subj
n_cond <- 2L
n_feat <- 140L
n_behav <- 2L

site_levels <- c("site_a", "site_b", "site_c")
site_subj <- rep(site_levels, each = n_site_subj)
condition_levels <- c("past", "future")

site_shift <- c(site_a = -1.0, site_b = 0.8, site_c = 0.2)
site_scale <- c(site_a = 0.9, site_b = 1.15, site_c = 1.0)

latent <- matrix(rnorm(n_subj * n_cond), nrow = n_subj, ncol = n_cond)
condition_shift <- matrix(rep(c(-0.8, 0.8), each = n_subj), nrow = n_subj, ncol = n_cond)
site_shift_mat <- matrix(site_shift[site_subj], nrow = n_subj, ncol = n_cond)
site_scale_mat <- matrix(site_scale[site_subj], nrow = n_subj, ncol = n_cond)

brain_raw <- matrix(rnorm(n_subj * n_cond * n_feat, sd = 0.7), nrow = n_subj * n_cond, ncol = n_feat)
brain_signal <- as.numeric(site_scale_mat * (1.6 * latent + 0.5 * condition_shift))
brain_raw[, 1:45] <- brain_raw[, 1:45] + brain_signal %o% rep(1.0, 45)
brain_raw[, 46:90] <- brain_raw[, 46:90] + brain_signal %o% rep(-0.9, 45)
brain_raw[, 91:110] <- brain_raw[, 91:110] + as.numeric(site_shift_mat) %o% rep(0.8, 20)

behav_raw <- cbind(
  rt = as.numeric(site_scale_mat * (1.3 * latent + 0.4 * condition_shift) + site_shift_mat + rnorm(n_subj * n_cond, sd = 0.35)),
  acc = as.numeric(site_scale_mat * (-1.1 * latent + 0.2 * condition_shift) - 0.6 * site_shift_mat + rnorm(n_subj * n_cond, sd = 0.35))
)

standardize_within_site <- function(x, site_obs) {
  out <- x
  for (site_name in unique(site_obs)) {
    idx <- which(site_obs == site_name)
    out[idx, ] <- scale(x[idx, , drop = FALSE])
  }
  out
}

site_obs <- rep(site_subj, times = n_cond)
brain_z <- standardize_within_site(brain_raw, site_obs)
behav_z <- standardize_within_site(behav_raw, site_obs)

site_summary <- do.call(
  rbind,
  lapply(site_levels, function(site_name) {
    idx <- which(site_obs == site_name)
    data.frame(
      site = site_name,
      n_subjects = sum(site_subj == site_name),
      n_observations = length(idx),
      rt_mean = mean(behav_z[idx, "rt"]),
      acc_mean = mean(behav_z[idx, "acc"]),
      stringsAsFactors = FALSE
    )
  })
)

stopifnot(
  nrow(site_summary) == 3L,
  all(abs(site_summary$rt_mean) < 1e-10),
  all(abs(site_summary$acc_mean) < 1e-10)
)
```

```{r site-summary}
site_summary
```

This example has three sites, `r n_subj` pooled subjects, and `r n_cond`
conditions. The hidden setup adds site-specific mean and scale shifts before
standardization, so the pooled fit has to recover the shared latent
brain-behavior relation rather than scanner/site offsets.

## How do you fit one pooled model and keep site information?

Use the usual builder path for behavioral PLS, then attach site labels before
running the fit.

```{r run-site-pooled-pls}
site_spec <- pls_spec() |>
  add_subjects(list(brain_z), groups = n_subj) |>
  add_conditions(condition_levels) |>
  add_behavior(behav_z) |>
  add_site_labels(site_subj) |>
  configure(method = "behavior", nperm = 120, nboot = 120)

site_result <- run(site_spec, progress = FALSE)
site_diag <- site_result$site_diagnostics
site_diag_infer <- site_pooling_diagnostics(
  site_result,
  spec = site_spec,
  infer = "score",
  nperm = 39,
  nboot = 79,
  progress = FALSE
)

stopifnot(
  inherits(site_result, "pls_result"),
  is.list(site_diag),
  is.list(site_diag_infer$score_heterogeneity),
  setequal(site_diag$sites, site_levels)
)
```

Check the pooled latent variables first, just as you would in an ordinary
behavior PLS run:

```{r pooled-significance}
pooled_summary <- cbind(
  pvalue = round(significance(site_result), 3),
  variance = round(singular_values(site_result, normalize = TRUE), 1)
)

stopifnot(
  pooled_summary[1, "pvalue"] <= 0.1,
  pooled_summary[1, "variance"] > pooled_summary[2, "variance"]
)

pooled_summary
```

LV1 is the pooled brain-behavior pattern. The important multisite question is
whether that pooled pattern looks similar across sites.

## What diagnostics do you get when site labels are present?

`site_pooling_diagnostics()` summarizes four things:

- sitewise score distributions
- sitewise brain-behavior score correlations
- site-specific refits compared with the pooled saliences
- leave-one-site-out reruns

Because `site` was provided in the builder, those diagnostics are already
attached to the result:

```{r site-correlation-summary}
site_corr_lv1 <- subset(site_diag$site_score_correlations, lv == 1)

stopifnot(
  nrow(site_corr_lv1) == length(site_levels),
  all(is.finite(site_corr_lv1$correlation)),
  length(unique(sign(site_corr_lv1$correlation))) == 1L,
  all(abs(site_corr_lv1$correlation) > 0.35)
)

site_corr_lv1
```

Every site shows the same direction of brain-behavior association on LV1. That
is the first thing you want to see in a pooled multisite analysis.

```{r plot-site-correlations, fig.cap = "Sitewise LV1 brain-behavior score correlations. All three sites show the same direction of association."}
ggplot2::ggplot(site_corr_lv1, ggplot2::aes(x = site, y = correlation, fill = site)) +
  ggplot2::geom_col(width = 0.7, show.legend = FALSE) +
  ggplot2::geom_hline(yintercept = 0, linetype = 2, color = "grey50") +
ggplot2::labs(x = "Site", y = "LV1 score correlation") +
  theme_pls()
```

## Can you test whether those site correlations are too heterogeneous?

Yes. `infer = "score"` adds subject-block bootstrap intervals for each site's
LV score correlation and a permutation p-value for cross-site heterogeneity of
those correlations. This is a better answer than "correlation below some fixed
cutoff" because it asks whether the site differences are larger than you would
expect from random subject-to-site reassignment under the pooled model.

```{r score-heterogeneity}
hetero_lv1 <- subset(site_diag_infer$score_heterogeneity$global_tests, lv == 1)
site_ci_lv1 <- subset(site_diag_infer$score_heterogeneity$site_intervals, lv == 1)

stopifnot(
  nrow(hetero_lv1) == 1L,
  hetero_lv1$perm_pvalue > 0.1,
  all(site_ci_lv1$conf_low <= site_ci_lv1$conf_high)
)

hetero_lv1
site_ci_lv1
```

In this synthetic example, LV1 remains highly consistent across sites, so the
heterogeneity p-value is not small. If that p-value were small, the pooled LV
could still be significant overall while failing the multisite stability check.

## Do the site-specific saliences resemble the pooled solution?

The site-specific reruns compare each site's own fit to the pooled fit using
cosine similarity in both feature space and design space.

```{r site-similarity-summary}
site_sim_lv1 <- subset(site_diag$site_fit_similarity, lv == 1)

stopifnot(
  nrow(site_sim_lv1) == length(site_levels),
  all(site_sim_lv1$feature_cosine > 0.6),
  all(site_sim_lv1$design_cosine > 0.6)
)

site_sim_lv1
```

Those similarities are not a replacement for interpretation, but they are a
useful quick check that the pooled LV is not being driven by one site alone.
If you want an inferential version of that check, use
`site_pooling_diagnostics(..., infer = "full")`, which adds site-label
permutation p-values for cumulative subspace concordance across the first `k`
LVs.

## What happens if you leave one site out?

Leave-one-site-out reruns are the strongest built-in stress test in the current
workflow. They refit the model on two sites, then project the held-out site
into that score space.

```{r leave-one-site-out}
loos_lv1 <- subset(site_diag$leave_one_site_out, lv == 1)

stopifnot(
  nrow(loos_lv1) == length(site_levels),
  all(is.finite(loos_lv1$heldout_score_correlation)),
  all(loos_lv1$feature_cosine > 0.5),
  all(loos_lv1$design_cosine > 0.5),
  length(unique(sign(loos_lv1$heldout_score_correlation))) == 1L
)

loos_lv1
```

If the held-out score correlations flip sign or collapse toward zero, the
pooled latent pattern is not really shared. In this synthetic example the
held-out site still lines up with the training fit, which is what you want in a
stable pooled analysis.

## When should you stop pooling and rethink the model?

Pooling is still a model choice. The diagnostics here are meant to tell you when
the pooled result is doing real shared work and when it is papering over site
heterogeneity.

Rethink the pooled model if you see any of these:

- a small score-heterogeneity p-value for the pooled LV
- sitewise LV correlations with different signs or non-overlapping sitewise bootstrap intervals
- low site-specific salience similarity to the pooled LV, especially if `infer = "full"` also shows weak subspace concordance
- leave-one-site-out fits that fail only when one particular site is held out
- score distributions that are dominated by one site rather than by the latent pattern

At that point you may want stronger harmonization, explicit site residualization,
or separate site-specific analyses.

## Where should you go next?

| Goal | Resource |
|:-----|:---------|
| Basic behavior PLS without site pooling | `vignette("behavior-pls")` |
| Scripted CLI and artifact workflow | `vignette("scripted-workflows")` |
| YAML contract for multisite pipelines | `vignette("pipeline-yaml-spec")` |
| Add site labels in the builder | `?add_site_labels` |
| Compute diagnostics from a fitted result | `?site_pooling_diagnostics` |
| Full behavior wrapper | `?behav_pls` |
