---
title: "Design-Subspace Summaries for Task PLS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Design-Subspace Summaries for Task PLS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)

ensure_local_plsrri <- function(required_exports) {
  pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."
  if (!file.exists(file.path(pkg_root, "DESCRIPTION")) ||
      !requireNamespace("devtools", quietly = TRUE)) {
    return(invisible(FALSE))
  }

  installed_exports <- if (requireNamespace("plsrri", quietly = TRUE)) {
    getNamespaceExports("plsrri")
  } else {
    character(0)
  }

  if (!all(required_exports %in% installed_exports)) {
    devtools::load_all(pkg_root, quiet = TRUE)
    return(invisible(TRUE))
  }

  invisible(FALSE)
}

ensure_local_plsrri(c(
  "pls_design",
  "design_cell_table",
  "decompose_design_terms",
  "test_design_terms",
  "compare_designs"
))
```

```{r load-package}
library(plsrri)
```

Task PLS gives you latent variables over a flattened set of group-by-condition
cells. When the conditions come from a factorial design, the next question is
often more specific: does the covariance pattern live mostly in a task effect,
a group-by-task interaction, or a higher-order term?

The design-subspace helpers answer that question on the design side of Task
PLS. They do not run voxelwise ANOVAs. They build factorial subspaces over the
same centered cell rows used by Task PLS and summarize how much cross-block
covariance lies in each subspace.

## What are the data?

This example uses synthetic data with two groups, two tasks, and two levels.
The planted signal includes a group-by-task effect, a level effect, and a
smaller three-way interaction.

```{r simulate-design, include = FALSE}
set.seed(20260509)

n_subjects <- 18
n_features <- 90

condition_key <- expand.grid(
  level = c("low", "high"),
  task = c("recog", "nback"),
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)
condition_key$condition <- paste(condition_key$task, condition_key$level, sep = "_")
condition_key <- condition_key[, c("condition", "task", "level")]

make_group_matrix <- function(group) {
  group_code <- if (identical(group, "sdam")) 1 else -1
  rows <- vector("list", n_subjects * nrow(condition_key))
  row_id <- 1L

  for (subject in seq_len(n_subjects)) {
    subject_offset <- rnorm(n_features, sd = 0.15)
    for (condition_id in seq_len(nrow(condition_key))) {
      task_code <- if (condition_key$task[[condition_id]] == "nback") 1 else -1
      level_code <- if (condition_key$level[[condition_id]] == "high") 1 else -1

      x <- rnorm(n_features, sd = 0.7) + subject_offset
      x[1:25] <- x[1:25] + 0.7 * group_code * task_code
      x[26:45] <- x[26:45] + 0.5 * level_code
      x[46:60] <- x[46:60] + 0.35 * group_code * task_code * level_code
      rows[[row_id]] <- x
      row_id <- row_id + 1L
    }
  }

  do.call(rbind, rows)
}

control <- make_group_matrix("control")
sdam <- make_group_matrix("sdam")
```

The condition key maps each PLS condition label back to the factors that
created it.

```{r show-condition-key}
condition_key
```

## How do we keep the Task PLS cross-block matrix?

Design-subspace summaries need the cell-level cross-block matrix used by Task
PLS before the singular value decomposition. Request it at fit time with
`keep_crossblock = TRUE`.

```{r fit-task-pls}
fit <- pls_spec() |>
  add_subjects(list(control, sdam), groups = c(n_subjects, n_subjects)) |>
  add_conditions(nrow(condition_key), labels = condition_key$condition) |>
  add_group_labels(c("control", "sdam")) |>
  configure(method = "task", meancentering = "grand_mean") |>
  run(progress = FALSE, keep_crossblock = TRUE)
```

```{r check-fit, include = FALSE}
stopifnot(
  inherits(fit, "pls_result"),
  is.list(fit$task_pls),
  is.matrix(fit$task_pls$crossblock),
  all(is.finite(fit$task_pls$crossblock))
)
```

The ordinary Task PLS result is still a latent-variable model. The new field is
only a compact summary object needed for design-subspace calculations; it does
not store the original subject-level imaging data.

## How do we define the factorial design?

Use `pls_design()` to describe the design over group-condition cells. The rows
come from the PLS result; the condition-level factors come from the condition
key.

```{r make-design}
design <- pls_design(
  ~ group * task * level,
  condition_key = condition_key,
  between = "group",
  within = c("task", "level")
)
```

You can inspect the cell table to confirm the row order. These are the rows of
the Task PLS cross-block matrix: all conditions for group 1, then all
conditions for group 2.

```{r show-cell-table}
design_cell_table(fit, design = design)[
  , c("row_index", "group", "condition", "task", "level")
]
```

```{r check-cell-table, include = FALSE}
cells <- design_cell_table(fit, design = design)
stopifnot(
  nrow(cells) == nrow(fit$task_pls$crossblock),
  identical(as.character(cells$condition[seq_len(nrow(condition_key))]), condition_key$condition)
)
```

## Which factorial terms does an LV resemble?

Start descriptively. `decompose_design_terms()` projects one design-side LV
onto the centered factorial subspaces and reports how much LV energy aligns
with each term. This helps interpret an already selected LV; it is not a
p-value.

```{r decompose-lv}
lv_terms <- decompose_design_terms(fit, lv = 1, design = design)
lv_terms$fraction <- round(lv_terms$fraction, 3)
lv_terms[order(lv_terms$fraction, decreasing = TRUE),
         c("term", "rank", "fraction")]
```

```{r check-lv-decomposition, include = FALSE}
stopifnot(
  all(is.finite(lv_terms$energy)),
  all(lv_terms$fraction >= 0),
  all(lv_terms$fraction <= 1 + 1e-8)
)
```

For this simulated example, LV1 should align strongly with the planted
factorial effects. The exact ordering is data-dependent because the example
still includes noise.

A plot of the same LV-level decomposition is useful when you want a quick
interpretive summary for a selected latent variable.

```{r plot-lv-terms, fig.cap = "Effect-coded decomposition of the LV1 design scores."}
plot_design_contrasts(fit, lv = 1, condition_key = condition_key)
```

## How much covariance is in each design subspace?

`test_design_terms()` works on the stored Task PLS cross-block matrix, not just
on one LV. The default statistic is `"trace"`, the total squared singular
value energy in each centered term subspace.

```{r test-terms}
term_tests <- test_design_terms(
  fit,
  design = design,
  statistic = "trace"
)
term_tests$share <- round(term_tests$statistic / sum(term_tests$statistic), 3)
term_tests$statistic <- round(term_tests$statistic, 2)
term_tests[, c("term", "rank", "statistic", "share", "p_value")]
```

```{r check-term-tests, include = FALSE}
stopifnot(
  all(is.finite(term_tests$statistic)),
  all(term_tests$rank >= 0),
  all(is.na(term_tests$p_value)),
  sum(term_tests$share) > 0.9
)
```

The `p_value` column is intentionally `NA` here. The current implementation
computes observed design-subspace summaries; permutation p-values for nested
factorial questions require a reduced-model-respecting null and are not
silently approximated.

If you want the most PLS-like single-root statistic instead of omnibus energy,
use `statistic = "largest_root"`.

```{r largest-root}
largest_root <- test_design_terms(
  fit,
  design = design,
  statistic = "largest_root"
)
largest_root$statistic <- round(largest_root$statistic, 2)
largest_root[, c("term", "rank", "statistic")]
```

```{r check-largest-root, include = FALSE}
trace_again <- test_design_terms(fit, design = design, statistic = "trace")
root_again <- test_design_terms(fit, design = design, statistic = "largest_root")
stopifnot(all(root_again$statistic <= trace_again$statistic + 1e-8))
```

## How do we ask a nested design question?

Use `compare_designs()` when your question is "what is added by the full model
beyond the reduced model?" The function residualizes the full design against
the reduced design in the same centered Task PLS row space.

```{r compare-designs}
task_beyond_group <- compare_designs(
  fit,
  design = design,
  reduced = ~ group,
  full = ~ group * task
)

interactions_beyond_main_effects <- compare_designs(
  fit,
  design = design,
  reduced = ~ group + task + level,
  full = ~ group * task * level
)

nested_tests <- rbind(task_beyond_group, interactions_beyond_main_effects)
nested_tests$statistic <- round(nested_tests$statistic, 2)
nested_tests[, c("reduced", "full", "added_terms", "rank", "statistic", "p_value")]
```

```{r check-nested-tests, include = FALSE}
stopifnot(
  all(is.finite(nested_tests$statistic)),
  all(nested_tests$rank >= 0),
  all(is.na(nested_tests$p_value))
)
```

The first comparison asks how much covariance is carried by task plus the
group-by-task interaction after accounting for group. The second asks how much
covariance is carried by all interactions after accounting for the three main
effects.

## How should the summaries be interpreted?

Keep three distinctions clear:

1. `plot_scores()` and `plot_design_contrasts()` help you interpret selected
   LVs.
2. `decompose_design_terms()` descriptively attributes one LV's design vector
   to factorial terms.
3. `test_design_terms()` and `compare_designs()` summarize covariance energy
   in centered design subspaces of the Task PLS cross-block matrix.

These are design-subspace summaries, not classical ANOVA tables and not
voxelwise tests. They preserve the multivariate PLS object while making the
factorial structure of the design explicit.

## Where to go next

Use `vignette("sdam-firstlevel-task-pls")` for a first-level image-map example
with design-score heatmaps and contrast plots. Use `?pls_design`,
`?decompose_design_terms`, `?test_design_terms`, and `?compare_designs` for the
function-level reference.
