---
title: "Design-Subspace Summaries for Task PLS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Design-Subspace Summaries for Task PLS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)

ensure_local_plsrri <- function(required_exports) {
  pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."
  if (!file.exists(file.path(pkg_root, "DESCRIPTION")) ||
      !requireNamespace("devtools", quietly = TRUE)) {
    return(invisible(FALSE))
  }

  installed_exports <- if (requireNamespace("plsrri", quietly = TRUE)) {
    getNamespaceExports("plsrri")
  } else {
    character(0)
  }

  if (!all(required_exports %in% installed_exports)) {
    devtools::load_all(pkg_root, quiet = TRUE)
    return(invisible(TRUE))
  }

  invisible(FALSE)
}

ensure_local_plsrri(c(
  "pls_design",
  "design_cell_table",
  "decompose_design_terms",
  "design_subspace_svd",
  "test_design_subspaces",
  "compare_design_subspaces"
))
```

```{r load-package}
library(plsrri)
```

Task PLS gives you latent variables over a flattened set of group-by-condition
cells. When the conditions come from a factorial design, the next question is
often more specific: does the covariance pattern live mostly in a task effect,
a group-by-task interaction, or a higher-order term?

The design-subspace helpers answer that question on the design side of Task
PLS. They do not run voxelwise ANOVAs. They build factorial subspaces over the
same centered cell rows used by Task PLS and summarize how much cross-block
covariance lies in each subspace.

## What are the data?

This example uses synthetic data with two groups, two tasks, and two levels.
The planted signal includes a group-by-task effect, a level effect, and a
smaller three-way interaction.

```{r simulate-design, include = FALSE}
set.seed(20260509)

n_subjects <- 18
n_features <- 90

condition_key <- expand.grid(
  level = c("low", "high"),
  task = c("recog", "nback"),
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)
condition_key$condition <- paste(condition_key$task, condition_key$level, sep = "_")
condition_key <- condition_key[, c("condition", "task", "level")]

make_group_matrix <- function(group) {
  group_code <- if (identical(group, "sdam")) 1 else -1
  rows <- vector("list", n_subjects * nrow(condition_key))
  row_id <- 1L

  for (subject in seq_len(n_subjects)) {
    subject_offset <- rnorm(n_features, sd = 0.15)
    for (condition_id in seq_len(nrow(condition_key))) {
      task_code <- if (condition_key$task[[condition_id]] == "nback") 1 else -1
      level_code <- if (condition_key$level[[condition_id]] == "high") 1 else -1

      x <- rnorm(n_features, sd = 0.7) + subject_offset
      x[1:25] <- x[1:25] + 0.7 * group_code * task_code
      x[26:45] <- x[26:45] + 0.5 * level_code
      x[46:60] <- x[46:60] + 0.35 * group_code * task_code * level_code
      rows[[row_id]] <- x
      row_id <- row_id + 1L
    }
  }

  do.call(rbind, rows)
}

control <- make_group_matrix("control")
sdam <- make_group_matrix("sdam")
```

The condition key maps each PLS condition label back to the factors that
created it.

```{r show-condition-key}
condition_key
```

## How do we keep the Task PLS cross-block matrix?

Design-subspace summaries need the cell-level cross-block matrix used by Task
PLS before the singular value decomposition. Request it at fit time with
`keep_crossblock = TRUE`.

```{r fit-task-pls}
spec <- pls_spec() |>
  add_subjects(list(control, sdam), groups = c(n_subjects, n_subjects)) |>
  add_conditions(nrow(condition_key), labels = condition_key$condition) |>
  add_group_labels(c("control", "sdam")) |>
  configure(method = "task", meancentering = "grand_mean")

fit <- run(spec, progress = FALSE, keep_crossblock = TRUE)
```

```{r check-fit, include = FALSE}
stopifnot(
  inherits(fit, "pls_result"),
  is.list(fit$task_pls),
  is.matrix(fit$task_pls$crossblock),
  all(is.finite(fit$task_pls$crossblock))
)
```

The ordinary Task PLS result is still a latent-variable model. The new field is
only a compact summary object needed for design-subspace calculations; it does
not store the original subject-level imaging data.

## How do we define the factorial design?

Use `pls_design()` to describe the design over group-condition cells. The rows
come from the PLS result; the condition-level factors come from the condition
key.

```{r make-design}
design <- pls_design(
  ~ group * task * level,
  condition_key = condition_key,
  between = "group",
  within = c("task", "level")
)
```

You can inspect the cell table to confirm the row order. These are the rows of
the Task PLS cross-block matrix: all conditions for group 1, then all
conditions for group 2.

```{r show-cell-table}
design_cell_table(fit, design = design)[
  , c("row_index", "group", "condition", "task", "level")
]
```

```{r check-cell-table, include = FALSE}
cells <- design_cell_table(fit, design = design)
stopifnot(
  nrow(cells) == nrow(fit$task_pls$crossblock),
  identical(as.character(cells$condition[seq_len(nrow(condition_key))]), condition_key$condition)
)
```

## Which factorial terms does an LV resemble?

Start descriptively. `decompose_design_terms()` projects one design-side LV
onto the centered factorial subspaces and reports how much LV energy aligns
with each term. This helps interpret an already selected LV; it is not a
p-value.

```{r decompose-lv}
lv_terms <- decompose_design_terms(fit, lv = 1, design = design)
lv_terms$fraction <- round(lv_terms$fraction, 3)
lv_terms[order(lv_terms$fraction, decreasing = TRUE),
         c("term", "rank", "fraction")]
```

```{r check-lv-decomposition, include = FALSE}
stopifnot(
  all(is.finite(lv_terms$energy)),
  all(lv_terms$fraction >= 0),
  all(lv_terms$fraction <= 1 + 1e-8)
)
```

For this simulated example, LV1 should align strongly with the planted
factorial effects. The exact ordering is data-dependent because the example
still includes noise.

A plot of the same LV-level decomposition is useful when you want a quick
interpretive summary for a selected latent variable.

```{r plot-lv-terms, fig.cap = "Effect-coded decomposition of the LV1 design scores."}
plot_design_contrasts(fit, lv = 1, condition_key = condition_key)
```

## How do we fit term-specific PLS components?

The ASCA-like step is `design_subspace_svd()`. It projects the stored Task PLS
cross-block matrix into each centered factorial subspace, then fits a separate
SVD for each term. This is a fitted term-specific PLS decomposition, not a
post-hoc label attached to a global LV.

```{r subspace-svd}
term_fit <- design_subspace_svd(
  fit,
  design = design,
  ncomp = 2,
  statistic = "trace"
)

term_fit$statistics[, c("term", "rank", "trace", "largest_root")]
```

```{r show-subspace-components}
components <- term_fit$components
components$percent_term_covariance <- round(components$percent_term_covariance, 3)
components[components$component == 1,
           c("term", "component", "singular_value", "percent_term_covariance")]
```

```{r check-subspace-svd, include = FALSE}
stopifnot(
  inherits(term_fit, "pls_design_subspace_svd"),
  all(is.finite(term_fit$statistics$trace)),
  all(term_fit$statistics$rank >= 0)
)
```

The term-level `trace` is the total squared singular-value energy in that
subspace. `largest_root` is the dominant PLS-like component for the same
projected matrix.

## Can we get resampling p-values?

Yes, against an explicitly named null. `test_design_subspaces()` recomputes
each subspace statistic over permuted Task PLS cross-block matrices and
returns a p-value per term.

The only null currently implemented is `permutation = "global_task_pls"` —
the same permutation scheme that ordinary Task PLS already uses for LV
significance. Concretely, for each permutation the routine

1. permutes condition labels within each subject and permutes subjects across
   groups (`pls_perm_order()`),
2. rebuilds the centered group-by-condition cross-block matrix from the
   permuted subject rows (`pls_get_covcor()`), and
3. recomputes each term-subspace statistic on the permuted cross-block.

Under this null **every factorial label is broken simultaneously** — there is
no group structure, no task structure, no interaction structure. The
question being asked for each term is therefore:

> Is the covariance projected into this term's subspace larger than what we
> would see if subject-condition labels were exchangeable?

That is the right null for an *omnibus* "is this term doing anything?"
question, and it is the standard PLS permutation null restricted to a
subspace. It is **not** a nested test: it cannot tell you whether
`group:task` adds anything *beyond* `group`, because the same permutation
that breaks `group:task` also breaks `group`. A reduced-model-preserving
null (Freedman–Lane–style) is the right tool for nested questions and is
filed as future work.

The `correction = "maxT"` option compares each observed term statistic to
the per-permutation maximum across all terms, giving family-wise adjusted
p-values across the table.

```{r test-terms}
set.seed(20260510)
term_tests <- test_design_subspaces(
  spec,
  fit = fit,
  design = design,
  statistic = "trace",
  nperm = 99,
  permutation = "global_task_pls",
  correction = "maxT"
)
term_tests$share <- round(term_tests$statistic / sum(term_tests$statistic), 3)
term_tests$statistic <- round(term_tests$statistic, 2)
term_tests[, c("term", "rank", "statistic", "share", "p_value", "p_adjusted")]
```

```{r check-term-tests, include = FALSE}
stopifnot(
  all(is.finite(term_tests$statistic)),
  all(term_tests$rank >= 0),
  all(is.finite(term_tests$p_value)),
  all(term_tests$p_value >= 0 & term_tests$p_value <= 1),
  sum(term_tests$share) > 0.9
)
```

If you want the most PLS-like single-root statistic instead of omnibus energy,
use `statistic = "largest_root"`.

```{r largest-root}
largest_root <- test_design_subspaces(
  fit,
  design = design,
  statistic = "largest_root"
)
largest_root$statistic <- round(largest_root$statistic, 2)
largest_root[, c("term", "rank", "statistic")]
```

```{r check-largest-root, include = FALSE}
trace_again <- test_design_subspaces(fit, design = design, statistic = "trace")
root_again <- test_design_subspaces(fit, design = design, statistic = "largest_root")
stopifnot(all(root_again$statistic <= trace_again$statistic + 1e-8))
```

## How do we ask a nested design question?

Use `compare_design_subspaces()` when your question is "what is added by the
full model beyond the reduced model?" The function residualizes the full design
against the reduced design in the same centered Task PLS row space, then fits
the delta-subspace SVD.

```{r compare-designs}
task_beyond_group <- compare_design_subspaces(
  spec,
  fit = fit,
  design = design,
  reduced = ~ group,
  full = ~ group * task,
  nperm = 99,
  permutation = "global_task_pls"
)

interactions_beyond_main_effects <- compare_design_subspaces(
  spec,
  fit = fit,
  design = design,
  reduced = ~ group + task + level,
  full = ~ group * task * level,
  nperm = 99,
  permutation = "global_task_pls"
)

nested_tests <- rbind(task_beyond_group, interactions_beyond_main_effects)
nested_tests$statistic <- round(nested_tests$statistic, 2)
nested_tests[, c("reduced", "full", "added_terms", "rank", "statistic", "p_value")]
```

```{r check-nested-tests, include = FALSE}
stopifnot(
  all(is.finite(nested_tests$statistic)),
  all(nested_tests$rank >= 0),
  all(is.finite(nested_tests$p_value))
)
```

The `statistic` column is the *observed* covariance energy in the
residualized delta subspace — i.e., the part of the full design that is
orthogonal to the reduced design after centering. This part of the answer
is purely descriptive and does not depend on the permutation null.

The `p_value` column, however, again uses the `"global_task_pls"` null.
That is, the delta statistic is compared to the distribution of delta
statistics computed on cross-block matrices where *all* subject-condition
labels have been permuted — not just the labels associated with the added
terms. So the p-value answers:

> Is the energy in the full-minus-reduced subspace larger than expected
> when there is no design structure at all?

That is informative and it is what classical Task PLS already gives you for
an LV. It is **not** the same as testing "does adding `task` and
`group:task` improve fit beyond `group` alone" in the strict nested sense:
a true nested test would permute under a null that preserves the reduced
model, and that null is not yet implemented. Use the descriptive
`statistic` and `rank` columns when you want to compare the *size* of the
added subspace, and treat the p-value as global-null evidence that
*something* in the delta subspace is non-trivial.

## How should the summaries be interpreted?

Keep four distinctions clear:

1. `plot_scores()` and `plot_design_contrasts()` help you interpret selected
   LVs.
2. `decompose_design_terms()` descriptively attributes one LV's design vector
   to factorial terms.
3. `design_subspace_svd()` fits term-specific PLS components from projected
   cross-block matrices.
4. `test_design_subspaces()` and `compare_design_subspaces()` add observed
   statistics and, when a `pls_spec` is available, global Task PLS permutation
   p-values.

These are design-subspace summaries, not classical ANOVA tables and not
voxelwise tests. They differ from classical ANOVA in three ways worth keeping
in mind:

- **The response is multivariate.** The statistic is squared singular-value
  energy of a projected cross-block matrix, not a univariate sum-of-squares.
  There is no F ratio and no residual mean square; the reference distribution
  is built by permutation, not by Gaussian-error theory.
- **The null is exchangeability of subject-condition labels.** It is the same
  null used for LV significance in ordinary Task PLS — labels are permuted,
  cross-blocks are rebuilt, statistics are recomputed. There is no
  independence or homoscedasticity assumption.
- **Term tests are global, not sequential.** The currently implemented null
  breaks all design structure at once. Each term's p-value answers "is this
  subspace doing anything?" against a no-design null, not "does this term
  add anything beyond the others?" The latter is what nested
  `compare_design_subspaces()` calls would ask once a reduced-model-preserving
  null is implemented.

Treated this way, the table is a multivariate, design-aware companion to the
LV table — not a substitute for ANOVA and not a voxelwise test.

## Where to go next

Use `vignette("sdam-firstlevel-task-pls")` for a first-level image-map example
with design-score heatmaps and contrast plots. Use `?pls_design`,
`?decompose_design_terms`, `?design_subspace_svd`,
`?test_design_subspaces`, and `?compare_design_subspaces` for the
function-level reference.
