---
title: "Within-Subject Seed PLS Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Within-Subject Seed PLS Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)

pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."
if (requireNamespace("plsrri", quietly = TRUE) &&
    !"ws_seed_pls" %in% getNamespaceExports("plsrri") &&
    requireNamespace("devtools", quietly = TRUE) &&
    file.exists(file.path(pkg_root, "DESCRIPTION"))) {
  devtools::load_all(pkg_root, quiet = TRUE)
}
```

```{r load-package}
library(plsrri)
```

Within-subject seed PLS is for a different question than ordinary `seed_pls()`.
If you care about whether a seed and the rest of the brain *co-fluctuate across
trials within each person*, you need trial-level data and a within-subject
connectivity measure. `ws_seed_pls()` builds those seed-to-voxel connectivity
maps first and only then runs task PLS on the resulting maps.

## Why not just use ordinary Seed PLS?

`seed_pls()` works on one seed value per subject-condition observation. It is
useful when you want across-subject covariance between seed activity and the
rest of the brain. `ws_seed_pls()` asks a different question: within a given
subject and condition, do the seed and voxel betas rise and fall together from
trial to trial?

```{r simulate-ws-data, include = FALSE}
set.seed(2026)

n_subj <- 16
n_trials <- 24
n_cond <- 2
n_vox <- 80
trials_per_cond <- n_trials / n_cond

condition_levels <- c("past", "future")

beta_lst <- vector("list", n_subj)
seed_lst <- vector("list", n_subj)
condition_lst <- vector("list", n_subj)

for (i in seq_len(n_subj)) {
  latent <- rnorm(n_trials)
  cond <- factor(rep(condition_levels, each = trials_per_cond), levels = condition_levels)

  seed <- latent + rnorm(n_trials, sd = 0.35)
  betas <- matrix(rnorm(n_trials * n_vox, sd = 0.7), nrow = n_trials, ncol = n_vox)

  past_idx <- which(cond == "past")
  future_idx <- which(cond == "future")

  betas[past_idx, 1:15] <- betas[past_idx, 1:15] + seed[past_idx] %o% rep(1.3, 15)
  betas[future_idx, 1:15] <- betas[future_idx, 1:15] - seed[future_idx] %o% rep(1.0, 15)
  betas[, 16:30] <- betas[, 16:30] + rep(c(-0.6, 0.6), each = trials_per_cond) %o% rep(1, 15)

  beta_lst[[i]] <- betas
  seed_lst[[i]] <- seed
  condition_lst[[i]] <- cond
}

ws_maps <- ws_seed_correlation(beta_lst, seed_lst, condition_lst, fisher_z = FALSE)

past_rows <- seq(1, n_subj * n_cond, by = 2)
future_rows <- seq(2, n_subj * n_cond, by = 2)

signal_corr <- c(
  past = mean(ws_maps$datamat[past_rows, 1:15], na.rm = TRUE),
  future = mean(ws_maps$datamat[future_rows, 1:15], na.rm = TRUE)
)

seed_means <- do.call(rbind, lapply(seq_len(n_subj), function(i) {
  data.frame(
    subject = i,
    condition = condition_levels,
    seed_mean = c(
      mean(seed_lst[[i]][condition_lst[[i]] == "past"]),
      mean(seed_lst[[i]][condition_lst[[i]] == "future"])
    )
  )
}))
seed_cond_diff <- with(seed_means, mean(seed_mean[condition == "future"]) - mean(seed_mean[condition == "past"]))

stopifnot(
  nrow(ws_maps$datamat) == n_subj * n_cond,
  ncol(ws_maps$datamat) == n_vox,
  is.finite(signal_corr[["past"]]),
  is.finite(signal_corr[["future"]]),
  signal_corr[["past"]] > 0.4,
  signal_corr[["future"]] < -0.3,
  abs(seed_cond_diff) < 0.35
)
```

This synthetic example was built so the seed has *similar mean activation* in the
two conditions, but its trial-by-trial coupling with voxels 1-15 flips sign:

```{r ws-rationale-check}
data.frame(
  quantity = c(
    "mean within-subject r in signal voxels (past)",
    "mean within-subject r in signal voxels (future)",
    "future - past difference in mean seed activation"
  ),
  value = round(c(signal_corr[["past"]], signal_corr[["future"]], seed_cond_diff), 2)
)
```

That is the raison d'etre of the ws-seed workflow: connectivity differences can
be strong even when the seed's average activation is not.

## What does the basic ws-seed workflow look like?

The basic workflow is:

1. compute one seed-to-voxel connectivity map per subject per condition
2. submit those maps to task PLS
3. inspect significance, design scores, and reliable voxels

```{r run-ws-seed}
ws_result <- ws_seed_pls(
  beta_lst = beta_lst,
  seed_lst = seed_lst,
  condition_lst = condition_lst,
  num_subj_lst = n_subj,
  fisher_z = FALSE,
  nperm = 200,
  nboot = 200,
  progress = FALSE
)

stopifnot(
  inherits(ws_result, "pls_result"),
  inherits(ws_result, "pls_task"),
  length(significance(ws_result)) >= 1L,
  all(is.finite(ws_result$s))
)
```

```{r ws-significance}
cbind(
  pvalue = round(significance(ws_result), 3),
  variance = round(singular_values(ws_result, normalize = TRUE), 1)
)
```

If the planted connectivity reversal is strong, LV1 should dominate:

```{r plot-ws-scree, fig.cap = "Singular values for the within-subject seed PLS example. LV1 captures the dominant condition-dependent connectivity pattern."}
plot_singular_values(ws_result)
```

The design scores should show opposite contributions from `past` and `future`:

```{r plot-ws-design-scores, fig.cap = "Design scores for LV1. The two conditions pull in opposite directions, reflecting the planted sign flip in within-subject connectivity."}
plot_scores(ws_result, lv = 1, type = "design", plot_type = "bar")
```

## Which voxels contribute reliably?

```{r ws-bsr}
ws_bsr <- bsr(ws_result, lv = 1)
reliable <- sum(abs(ws_bsr) > 2)

stopifnot(
  all(is.finite(ws_bsr)),
  reliable > 10L
)
```

`r reliable` of `r n_features(ws_result)` voxels pass the $|BSR| > 2$
threshold:

```{r plot-ws-bsr, echo = FALSE, fig.cap = "Bootstrap ratio profile for LV1. Voxels 1-15 carry the planted within-subject connectivity effect.", fig.height = 3}
plot(as.numeric(ws_bsr), type = "h", col = "steelblue",
     xlab = "Voxel", ylab = "Bootstrap Ratio (LV1)",
     main = NULL, las = 1)
abline(h = c(-2, 2), lty = 2, col = "grey50")
```

## How do multiple seeds work?

The general multiseed layout is `seed_condition`:

- rows = `subject × seed × condition`
- columns = voxels

That makes rotated and non-rotated ws-seed analyses use the same underlying
representation.

```{r simulate-multiseed, include = FALSE}
set.seed(2027)

multi_seed_lst <- vector("list", n_subj)
multi_beta_lst <- vector("list", n_subj)
multi_condition_lst <- vector("list", n_subj)

for (i in seq_len(n_subj)) {
  cond <- factor(rep(condition_levels, each = trials_per_cond), levels = condition_levels)
  latent_a <- rnorm(n_trials)
  latent_b <- rnorm(n_trials)

  seeds <- cbind(
    Precuneus = latent_a + rnorm(n_trials, sd = 0.35),
    IPS = latent_b + rnorm(n_trials, sd = 0.35)
  )

  betas <- matrix(rnorm(n_trials * n_vox, sd = 0.75), nrow = n_trials, ncol = n_vox)
  past_idx <- which(cond == "past")
  future_idx <- which(cond == "future")

  betas[past_idx, 1:12] <- betas[past_idx, 1:12] + seeds[past_idx, "Precuneus"] %o% rep(1.0, 12)
  betas[future_idx, 1:12] <- betas[future_idx, 1:12] - seeds[future_idx, "Precuneus"] %o% rep(0.8, 12)
  betas[past_idx, 13:24] <- betas[past_idx, 13:24] - seeds[past_idx, "IPS"] %o% rep(0.9, 12)
  betas[future_idx, 13:24] <- betas[future_idx, 13:24] + seeds[future_idx, "IPS"] %o% rep(1.1, 12)

  multi_seed_lst[[i]] <- seeds
  multi_beta_lst[[i]] <- betas
  multi_condition_lst[[i]] <- cond
}

multi_result <- ws_seed_pls(
  beta_lst = multi_beta_lst,
  seed_lst = multi_seed_lst,
  condition_lst = multi_condition_lst,
  num_subj_lst = n_subj,
  seed_labels = c("Precuneus", "IPS"),
  nonrotated = TRUE,
  fisher_z = FALSE,
  nperm = 0,
  nboot = 0,
  progress = FALSE
)

stopifnot(
  multi_result$num_cond == 4L,
  identical(multi_result$conditions, c("Precuneus:past", "Precuneus:future", "IPS:past", "IPS:future")),
  identical(multi_result$ws_seed_info$layout, "seed_condition"),
  identical(multi_result$feature_layout$kind, "voxel_map")
)
```

```{r multiseed-layout-check}
data.frame(
  condition_cell = multi_result$conditions,
  design_row = seq_along(multi_result$conditions)
)
```

This row expansion is what makes a targeted non-rotated design meaningful: you
can now test specific seed-by-condition hypotheses across explicit cells instead
of hiding seeds in the feature dimension.

## When would you use `stacked_seed_features` instead?

`stacked_seed_features` keeps the original task-PLS row structure:

- rows = `subject × condition`
- columns = `voxel × seed`

That can still be useful, but it answers a different question because the seed
dimension is now part of the feature space rather than the condition space.

```{r stacked-layout}
stacked_result <- ws_seed_pls(
  beta_lst = multi_beta_lst,
  seed_lst = multi_seed_lst,
  condition_lst = multi_condition_lst,
  num_subj_lst = n_subj,
  seed_labels = c("Precuneus", "IPS"),
  layout = "stacked_seed_features",
  fisher_z = FALSE,
  nperm = 0,
  nboot = 0,
  progress = FALSE
)

stopifnot(
  stacked_result$num_cond == 2L,
  identical(stacked_result$ws_seed_info$layout, "stacked_seed_features"),
  identical(stacked_result$feature_layout$kind, "voxel_seed"),
  nrow(stacked_result$u) == n_vox * 2L
)
```

```{r stacked-layout-check}
data.frame(
  property = c("num_cond", "feature layout", "features in u"),
  value = c(
    stacked_result$num_cond,
    stacked_result$feature_layout$kind,
    nrow(stacked_result$u)
  )
)
```

The general recommendation is:

- use `seed_condition` when seed identity should behave like an experimental
  cell
- use `stacked_seed_features` only when you explicitly want the seed dimension
  folded into the feature space

## Where to go next

| Goal | Resource |
|:-----|:---------|
| Task PLS workflow | `vignette("plsrri")` |
| Across-subject seed workflow | `vignette("multiblock-and-seed")` |
| Scripted API and CLI workflow | `vignette("scripted-workflows")` |
| Core helper for trial maps | `?ws_seed_correlation` |
| Main wrapper | `?ws_seed_pls` |
| Trial-level builder entry point | `?add_trial_data` |
